---
title: "SIMPER"
author: " "
date: "2024-05-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(vegan)

# Stop using scientific notation for numbers
options(scipen = 999)
```

# Approach

To assess which diet items contributed the most to observed differences in coyote diets among territories, we used similarity percentage analysis based on the dissimilarity matrix of the RRA data ("simper" in vegan).

# Prepare data

Prepare a diet data frame where each row is a territory and each column is the average RRA of each diet item.

Load diet data and metadata:

```{r}
# Load diet data
sp.RRA <- read.csv("Filtering-and-QC/sp.RRA.clean.csv") %>% select(-c(X))

# Make sample list column
sample_list <- sp.RRA %>% select(SampleID)

# Load metadata, add territories, seasons, filter out Stern Grove and Sunset Reservoir, and select relevant columns
metadata <- read.csv("Filtering-and-QC/All_Metadata.csv") %>% 
  
  # Add territories
  mutate(Territory = case_when(
    Site == "BERN" | Site == "STMA" ~ "Bernal",
    Site == "GLEN" | Site == "LAGU" | Site == "TWIN" | Site == "MTDA" | Site == "GGHP" ~ "Glen",
    Site == "SFGC" | Site == "OLYM" | Site == "LAKE" | Site == "FTFU" ~ "Merced",
    Site == "BVP" | Site == "CHP" ~ "Corona",
    Site == "TGPH" | Site == "HYDE" | Site == "FORT" ~ "Coit",
    Site == "GGPE" ~ "GGP - East",
    Site == "GGPW" | Site == "GGP" ~ "GGP - West",
    Site == "LEND" | Site == "LAFY" ~ "Land's End",
    Site == "MCLA" ~ "McLaren",
    Site == "PRES" ~ "Presidio",
    Site == "STFA" ~ "St. Francis",
    Site == "SUNS" ~ "Sunset Res",
    Site == "STGV" ~ "Stern Grove")) %>% 
  
  # Remove territories with low sample size
  filter(Territory != "Sunset Res" & Territory != "Stern Grove") %>%
  
  # Select only relevant columns
  select(c(SampleID, Territory)) %>%
  
  # Only keep metadata for samples in the final diet data frame
  inner_join(sample_list, by = "SampleID")
```

Prepare diet data:

```{r}
# Filter out samples from diet matrix that are not in the metadata 
sp.RRA.filt <- left_join(metadata, sp.RRA, by = "SampleID") %>% 
  select(-c(Territory))

territory.sp.RRA <- sp.RRA.filt %>% 
  
  # Pivot data frame
  pivot_longer(-SampleID, names_to="Species", values_to = "RRA") %>% 
  
  # Join metadata to diet data
  left_join(metadata, by = "SampleID") %>% 
  
  # Group by territory and species and calculate mean RRA
  group_by(Territory, Species) %>% 
  summarize(meanRRA = 100*mean(RRA), .groups="drop") %>% 
  
  # Pivot back to wide format
  pivot_wider(names_from = "Species", values_from = "meanRRA") %>% 
  
  # Remove diet items with 0 detections across all scats (diet items in dropped territories)
  select_if(~ !is.numeric(.) || sum(.) != 0)

# Make territories the row names
names <- territory.sp.RRA %>% select(Territory)

territory.sp.RRA <- territory.sp.RRA %>% # remove territory column
  select(-c(Territory))

rownames(territory.sp.RRA) <- names$Territory # add as row names
```

# Run SIMPER

If group argument is missing or has only one level, contributions are estimated for non-grouped data and dissimilarities show the overall heterogeneity in species abundances.

```{r}
# Run simper analysis (with no group)
territory.simper <- simper(territory.sp.RRA, permutations = 999, ordered = TRUE)

# Pull out overall dissimilarity
as.numeric(territory.simper$total["overall"])*100

# Make data frame with final table including species contribution to overall dissimilarity, standard deviation of contribution, percent contribution, and cumulative contribution
output <- summary(territory.simper)["total"][[1]] %>% 
  rownames_to_column("Species") %>%
  select(-c(ratio)) %>% 
  mutate(perc_cont = average/sum(average)) %>% # add percent contribution column
  relocate(perc_cont, .before = cumsum) %>% 
  mutate_if(is.numeric, ~ . * 100) %>%
  rename

output$Species <- gsub('\\.', ' ', output$Species) # change periods to spaces

colnames(output) <- c("Diet Item", "Average Contribution to Overall Dissimilarity", "SD of Contribution", "Percent Contribution", "Cumulative Contribution") # change column names

print(output)
```
