---
title: "Regression-Analyses"
author: "Tali Caspi"
date: "2024-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set custom plot theme
theme_custom <- function() {
  theme_classic() +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        strip.background = element_rect(fill = "grey80", color = NA),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        axis.title.y = element_text(size=12),
        axis.title.x = element_text(size=12))}

# Define function for standard error of proportion
se.prop <- function(detections, total.scats) {
  p_hat <- detections / total.scats
  standard_error <- sqrt((p_hat * (1 - p_hat)) / total.scats)
  return(standard_error)
}

# Define function for turning RRA to FOO
convert.FOO <- function(x) ifelse(x >0, 1, 0)

# Load packages
library(tidyverse)
library(plotrix)
library(mgcv)
library(marginaleffects)
library(DHARMa)
library(car)
library(ggpubr)
```

# Anthropogenic food x impervious surface cover

## Data preparation

Load data:

```{r}
# Load diet data
func.RRA <- read.csv("Filtering-and-QC/func.RRA.clean.csv") %>% select(-c(X))
func.FOO <- read.csv("Filtering-and-QC/func.FOO.clean.csv") %>% select(-c(X))

# Load metadata
metadata <- read.csv("Filtering-and-QC/All_Metadata.csv")

# Load covariate data
covariates <- read.csv("Regression-Analyses/Territory_All_Covariates.csv")
```

Prepare territory metadata:

```{r}
# Generate territory IDs from site data
territories <- metadata %>% 
  select(SampleID, Site) %>%  
  mutate(Territory = case_when(
    Site == "BERN" | Site == "STMA" ~ "Bernal",
    Site == "GLEN" | Site == "LAGU" | Site == "TWIN" | Site == "MTDA" | Site == "GGHP" ~ "Glen",
    Site == "SFGC" | Site == "OLYM" | Site == "LAKE" | Site == "FTFU" ~ "Merced",
    Site == "BVP" | Site == "CHP" ~ "Corona",
    Site == "TGPH" | Site == "HYDE" | Site == "FORT" ~ "Coit",
    Site == "GGPE" ~ "GGP - East",
    Site == "GGPW" | Site == "GGP" ~ "GGP - West",
    Site == "LEND" | Site == "LAFY" ~ "Land's End",
    Site == "MCLA" ~ "McLaren",
    Site == "PRES" ~ "Presidio",
    Site == "STFA" ~ "St. Francis",
    Site == "SUNS" ~ "Sunset Res",
    Site == "STGV" ~ "Stern Grove"))

# Make character vector of territories to drop (low sample size)
drop.sites <- c("Stern Grove", "Sunset Res")
```

Merge territories and covariate data to diet data:

```{r}
func.RRA.cov <- left_join(func.RRA, territories, by = "SampleID") %>%
  left_join(covariates, by = "Territory")

func.FOO.cov <- left_join(func.FOO, territories, by = "SampleID") %>%
  left_join(covariates, by = "Territory")
```

## Calculate territory-level RRA and FOO

Calculate mean RRA and overall FOO for anthropogenic foods for each territory with standard errors:

```{r}
# Define input data frames
clean.inputs.RRA <- list(func.RRA.cov)
clean.inputs.FOO <- list(func.FOO.cov)

# Define function
calc.anthro.RRA <- function(x) {x %>% 
  filter(!Territory %in% drop.sites) %>% 
  group_by(Territory) %>% 
  summarize(meanRRA = mean(Anthropogenic), 
            se = std.error(Anthropogenic),
            meanUrban = round(mean(Urban), digits=3), 
            meanISA = round(mean(ISA), digits=3),
            meanFood.Services = mean(Food.Services),
            meanPop.Den = mean(Pop.Den.2020),
            meanHouse.Den = mean(Housing.Den.2020))}

calc.anthro.FOO <- function(x) {x %>%
  filter(!Territory %in% drop.sites) %>% 
  group_by(Territory) %>% 
  summarize(detections = sum(Anthropogenic),
            total.scats=n(), 
            FOO=(detections/total.scats), 
            se = se.prop(detections,total.scats),
            meanUrban=round(mean(Urban), digits=3),
            meanISA = round(mean(ISA), digits=3),
            meanFood.Services = mean(Food.Services),
            meanPop.Den = mean(Pop.Den.2020),
            meanHouse.Den = mean(Housing.Den.2020))}

# Apply function
results.anthro.RRA <- lapply(clean.inputs.RRA, calc.anthro.RRA)
results.anthro.FOO <- lapply(clean.inputs.FOO, calc.anthro.FOO)

# Save the output
anthro.RRA <- results.anthro.RRA[[1]]
anthro.FOO <- results.anthro.FOO[[1]]
```

## Build models

We fit a beta regression for RRA (function "gam" with family betar in mgcv) and a quasibinomial GLM (function "glm" in stats) for FOO to test the effect of percent cover of impervious surfaces, a measure of urbanization, on the proportion and frequency of anthropogenic food in each territory.

### Relative read abundance

```{r}
# Construct model and interpret model
mod.RRA.ISA <- gam(meanRRA ~ meanISA, family=betar(link="logit"), data = anthro.RRA)

summary(mod.RRA.ISA)

# Calculate deviance explained
1 - mod.RRA.ISA$deviance/mod.RRA.ISA$null.deviance

# Calculate average marginal effects
mod.RRA.ISA |> avg_slopes()

# Check model fit
sim.RRA.ISA <- simulateResiduals(mod.RRA.ISA)
plot(sim.RRA.ISA)
```

Make predictions:

```{r}
# Generate new ISA data for making predictions
new_data <- data.frame(meanISA = seq(0, 100, by=0.1))

# Make predictions on the link scale, calculate confidence interval boundaries on the link scale, then back transform using the inverse link to get boundaries on the response scale
gam_preds <- as.data.frame(mgcv::predict.gam(mod.RRA.ISA, newdata = new_data, se.fit = TRUE, type="link")) %>% 
  mutate(meanISA = new_data$meanISA, .before="fit") %>% 
  mutate(upr = fit + (1.96 * se.fit),
         lwr = fit - (1.96 * se.fit),
         meanRRA = mod.RRA.ISA$family$linkinv(fit),
         upr_adj = mod.RRA.ISA$family$linkinv(upr),
         lwr_adj = mod.RRA.ISA$family$linkinv(lwr))
```

### Frequency of occurrence

```{r}
# Construct and interpret model
mod.FOO.ISA <- glm(FOO ~ meanISA, family = "quasibinomial", weights = anthro.FOO$total.scats, data=anthro.FOO)

summary(mod.FOO.ISA)

# Calculate deviance explained
1 - mod.FOO.ISA$deviance/mod.FOO.ISA$null.deviance

# Calculate average marginal effects
mod.FOO.ISA |> avg_slopes()

# Check that residuals are normally distributed
hist(resid(mod.FOO.ISA))

# Check that residual plots are evenly dispersed around 0
plot(resid(mod.FOO.ISA)~fitted(mod.FOO.ISA)) + abline(h=0)

# Check that in the qqplot residuals fall along the blue line
qqPlot(resid(mod.FOO.ISA))

# Check for homogeneity of variance
plot(mod.FOO.ISA)
```

Make predictions:

```{r}
# Generate new ISA data for making predictions
new_data_FOO <- data.frame(meanISA = seq(0, 100, by=0.1))

# Make predictions
FOO_preds <- as.data.frame(predict(mod.FOO.ISA, newdata = new_data_FOO, se.fit = TRUE, type="link")) %>% 
  mutate(meanISA = new_data_FOO$meanISA, .before="fit") %>% 
  mutate(upr = fit + (1.96 * se.fit),
         lwr = fit - (1.96 * se.fit),
         FOO = mod.FOO.ISA$family$linkinv(fit),
         upr_adj = mod.FOO.ISA$family$linkinv(upr),
         lwr_adj = mod.FOO.ISA$family$linkinv(lwr))
```

## Plot

```{r}
p.anthropogenic <- ggplot()+
  
  # RRA data - Beta regression
  geom_smooth(data = gam_preds, aes(x = meanISA, y = meanRRA,
                                    ymin = lwr_adj, ymax = upr_adj), stat = 'identity',
              color="#1F6481", fill="#1F6481", alpha=0.5)+
  geom_point(data=anthro.RRA, aes(x=meanISA, y=meanRRA),
             size=1.5, color="#044B69", alpha=0.8)+
  geom_errorbar(data=anthro.RRA,
                aes(x=meanISA, y=meanRRA,
                    ymin=meanRRA-se, ymax=meanRRA+se), 
                width=0.5, color="#044B69", alpha=0.8)+

  # FOO data - Quasibinomial GLM
  geom_smooth(data=FOO_preds, aes(x = meanISA, y = FOO,
                                  ymin = lwr_adj, ymax = upr_adj), stat = 'identity',
              color="#FAC5A6", fill="#FAC5A6", alpha=0.5)+ 
  geom_errorbar(data=anthro.FOO,
                aes(x=meanISA, y=FOO,
                    ymin=FOO-se, ymax=FOO+se), 
                width=0.5, color="#ECA276", alpha=0.8)+
  geom_point(data=anthro.FOO, aes(x=meanISA, y=FOO),
             size=1.5, color="#ECA276", alpha=0.8)+
  
# Graphics
  labs(x = "Impervious Surface Cover",
       y = "Anthropogenic Food (FOO or RRA ± SE)") +
  scale_y_continuous(limits=c(0,1),
                     breaks=c(0, 0.25, 0.50, 0.75, 1),
                     labels=c("0","25", "50", "75", "100"))+
  scale_x_continuous(expand=c(0,0), limits=c(0,102))+
  theme_custom() 

p.anthropogenic
```

# Invasive human-commensal rodents x number of food services

## Data preparation

Load data:

```{r}
sp.RRA <- read.csv("Filtering-and-QC/sp.RRA.clean.csv") %>% select(-c(X))
sp.FOO <- read.csv("Filtering-and-QC/sp.FOO.clean.csv") %>% select(-c(X))
```

Merge territories to diet data:

```{r}
sp.RRA.site <- left_join(sp.RRA, territories, by = "SampleID") %>% 
  left_join(covariates, by = "Territory")

sp.FOO.site <- left_join(sp.FOO, territories, by = "SampleID") %>% 
  left_join(covariates, by = "Territory")
```

## Calculate territory-level RRA and FOO for urban rodents

```{r}
# Combine Norway rat, Black rat, and House mouse
urban.rats.mice <- c("Rattus.norvegicus", "Rattus.rattus", "Mus.musculus")

# Add column that sums up the RRA for all urban rats and mice
combine.rodents.RRA <- sp.RRA.site %>% 
  mutate(urban.rodents = rowSums(select(., c(urban.rats.mice)))) %>% 
  relocate(urban.rodents, .after=SampleID) %>% 
  select(-c(Acipenser:Zalophus.californianus))

combine.rodents.FOO <- sp.FOO.site %>% 
  mutate(urban.rodents = Rattus.norvegicus + Rattus.rattus + Mus.musculus) %>% 
  mutate_at("urban.rodents", convert.FOO) %>% 
  relocate(urban.rodents, .after=SampleID) %>% 
  select(-c(Acipenser:Zalophus.californianus))

# Define input data frames
inputs.RRA <- list(combine.rodents.RRA)
inputs.FOO <- list(combine.rodents.FOO)

# Define function
calc.rodents.RRA <- function(x) {x %>% 
  filter(!Territory %in% drop.sites) %>% 
  group_by(Territory) %>% 
  summarize(meanRRA = mean(urban.rodents),
            se = std.error(urban.rodents),
            meanUrban = round(mean(Urban), digits=3), 
            meanISA = round(mean(ISA), digits=3),
            meanFood.Services = mean(Food.Services),
            meanPop.Den = mean(Pop.Den.2020),
            meanHouse.Den = mean(Housing.Den.2020))}

calc.rodents.FOO <- function(x) {x %>% 
  filter(!Territory %in% drop.sites) %>% 
  group_by(Territory) %>% 
  summarize(detections = sum(urban.rodents), 
            total.scats=n(), 
            FOO=detections/total.scats,
            se = se.prop(detections,total.scats),
            meanUrban = round(mean(Urban), digits=3), 
            meanISA = round(mean(ISA), digits=3),
            meanFood.Services = mean(Food.Services),
            meanPop.Den = mean(Pop.Den.2020),
            meanHouse.Den = mean(Housing.Den.2020))}

# Apply function
results.rodents.RRA <- lapply(inputs.RRA, calc.rodents.RRA)
results.rodents.FOO <- lapply(inputs.FOO, calc.rodents.FOO)

# Save the output as four separate dataframes
rodents.RRA <- results.rodents.RRA[[1]]
rodents.FOO <- results.rodents.FOO[[1]]
```

## Build models

We fit a beta regression for RRA (function "gam" with family betar in mgcv) and a quasibinomial GLM (function "glm" in stats) for FOO to test the effect of percent cover of impervious surfaces, a measure of urbanization, on the proportion and frequency of anthropogenic food in each territory.

### Relative read abundance

```{r}
# Construct model and interpret model
mod.RRA.rest <- gam(meanRRA ~ meanFood.Services, family=betar(link="logit"), data = rodents.RRA)

summary(mod.RRA.rest)

# Calculate deviance explained
1 - mod.RRA.rest$deviance/mod.RRA.rest$null.deviance

# Calculate average marginal effects
mod.RRA.rest|> avg_slopes() 

# Check model fit
sim.RRA.rest <- simulateResiduals(mod.RRA.rest)
plot(sim.RRA.rest)
```

Make predictions:

```{r}
# Generate new food services data for making predictions
new_data <- data.frame(meanFood.Services = seq(0, 350, by=0.1))

# Make predictions on the link scale, calculate confidence interval boundaries on the link scale, then back transform using the inverse link to get boundaries on the response scale
gam_preds <- as.data.frame(mgcv::predict.gam(mod.RRA.rest, newdata = new_data, se.fit = TRUE, type="link")) %>% 
  mutate(meanFood.Services = new_data$meanFood.Services, .before="fit") %>% 
  mutate(upr = fit + (1.96 * se.fit),
         lwr = fit - (1.96 * se.fit),
         meanRRA = mod.RRA.rest$family$linkinv(fit),
         upr_adj = mod.RRA.rest$family$linkinv(upr),
         lwr_adj = mod.RRA.rest$family$linkinv(lwr))
```

### Frequency of occurrence

```{r}
# Construct and interpret model
mod.FOO.rest <- glm(FOO~meanFood.Services, family=quasibinomial(link="logit"), weights = total.scats, data=rodents.FOO)

summary(mod.FOO.rest)

# Calculate deviance explained
1 - mod.FOO.rest$deviance/mod.FOO.rest$null.deviance

# Calculate average marginal effects
mod.FOO.rest|> avg_slopes()

# Check that residuals are normally distributed
hist(resid(mod.FOO.rest))

# Check that residual plots are evenly dispersed around 0
plot(resid(mod.FOO.rest)~fitted(mod.FOO.rest)) + abline(h=0)

# Check that in the qqplot residuals fall along the blue line
qqPlot(resid(mod.FOO.rest))

# Check for homogeneity of variance
plot(mod.FOO.rest)
```

Make predictions:

```{r}
# Generate new ISA data for making predictions
new_data_FOO <- data.frame(meanISA = seq(0, 100, by=0.1))

# Make predictions
FOO_preds <- as.data.frame(predict(mod.FOO.rest, newdata = new_data, se.fit = TRUE, type="link")) %>% 
  mutate(meanFood.Services = new_data$meanFood.Services, .before="fit") %>% 
  mutate(upr = fit + (1.96 * se.fit),
         lwr = fit - (1.96 * se.fit),
         FOO = mod.FOO.rest$family$linkinv(fit),
         upr_adj = mod.FOO.rest$family$linkinv(upr),
         lwr_adj = mod.FOO.rest$family$linkinv(lwr))
```

## Plot

```{r}
p.rodents <- ggplot()+
  
  # RRA data - GAM model with betar
  geom_smooth(data = gam_preds, aes(x = meanFood.Services, y = meanRRA,
                                    ymin = lwr_adj, ymax = upr_adj), stat = 'identity',
              color="#1F6481", fill="#1F6481", alpha=0.5)+
  geom_point(data=rodents.RRA, aes(x=meanFood.Services, y=meanRRA),
             size=1.5, color="#044B69", alpha=0.8)+
  geom_errorbar(data=rodents.RRA,
                aes(x=meanFood.Services, y=meanRRA,
                    ymin=meanRRA-se, ymax=meanRRA+se), 
                width=1.6, color="#044B69", alpha=0.8)+

  # FOO data - Quasibinomial GLM
  geom_smooth(data=FOO_preds, aes(x = meanFood.Services, y = FOO,
                                  ymin = lwr_adj, ymax = upr_adj), stat = 'identity',
              color="#FAC5A6", fill="#FAC5A6", alpha=0.5)+ 
  geom_errorbar(data=rodents.FOO,
                aes(x=meanFood.Services, y=FOO,
                    ymin=FOO-se, ymax=FOO+se), 
                width=1.6, color="#ECA276", alpha=0.8)+
  geom_point(data=rodents.FOO, aes(x=meanFood.Services, y=FOO),
             size=1.5, color="#ECA276", alpha=0.8)+
  
# Graphics
  labs(x = "Number of Food Services",
       y = "Nuisance Rodents (FOO or RRA ± SE)") +
  scale_y_continuous(limits=c(0,1),
                     breaks=c(0, 0.25, 0.50, 0.75, 1),
                     labels=c("0","25", "50", "75", "100"))+
  scale_x_continuous(expand=c(0,0), limits=c(0,350))+
  theme_custom()

p.rodents
```

# Combine plots

```{r}
ggarrange(p.anthropogenic, p.rodents,
          labels = c("(a)", "(b)"),
          hjust=-2.8,
          align="hv")
```
