---
title: "iNEXT"
author: 
date: "2024-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(iNEXT)
library(data.table)
library(tidyverse)
library(pals)
```

# Approach

We calculated and compared sample coverage, species richness, Hill-Shannon diversity, and Hill-Simpson diversity for territories and individuals on the basis of sample incidence data with the R package iNEXT. Sample coverage is the proportion of individuals (or in this case diet items) in an assemblage that belong to species represented in the sample. The three metrics of diversity differ in how they scale rarity: species richness is the most sensitive and Hill-Simpson the least sensitive to rare species. 

We calculated dietary rarefaction curves with iNEXT to determine if samples within territories and individuals reached a species diversity asymptote, indicating sample sizes large enough for comparison. We were unable to obtain an equal number of scat samples per individual. Therefore, we limited our individual-level analyses to the coyotes that were sampled the most frequently and with high estimated coverage, indicating that most of their diet items were represented in their samples (≥8 scats; code below). 

We used genetic data and local observations to assign frequently samples individuals to family groups. If the majority of an individual’s scats were collected in one territory over the course of the study or the individual was observed repeatedly in a territory, we felt confident in assigning the individual to the family group associated with that territory. One frequently sampled individual was designated as transient given inconsistency in where his fecal samples were collected.

# Prepare and format data

Following the format for the incidence_raw data structure for the iNEXT package.

## Territories

```{r}
# Loading species dataframes of the combined data
sp.FOO <- read.csv("Filtering-and-QC/sp.FOO.clean.csv") %>% select(-c(X))

# Load metadata, add territory, and select columns
metadata <- read.csv("Filtering-and-QC/All_Metadata.csv") %>% 
  mutate(Territory = case_when(
    Site == "BERN" | Site == "STMA" ~ "Bernal",
    Site == "GLEN" | Site == "LAGU" | Site == "TWIN" | Site == "MTDA" | Site == "GGHP" ~ "Glen",
    Site == "SFGC" | Site == "OLYM" | Site == "LAKE" | Site == "FTFU" ~ "Merced",
    Site == "BVP" | Site == "CHP" ~ "Corona",
    Site == "TGPH" | Site == "HYDE" | Site == "FORT" ~ "Coit",
    Site == "GGPE" ~ "GGP - East",
    Site == "GGPW" | Site == "GGP" ~ "GGP - West",
    Site == "LEND" | Site == "LAFY" ~ "Land's End",
    Site == "MCLA" ~ "McLaren",
    Site == "PRES" ~ "Presidio",
    Site == "STFA" ~ "St. Francis",
    Site == "SUNS" ~ "Sunset Res",
    Site == "STGV" ~ "Stern Grove")) %>% 
  select(c(SampleID, Category, Individual, Territory))

# Create character vector of sites to drop
drop.sites <- c("Sunset Res", "Stern Grove")

# Merge data frames and add territory data
sp_metadata <- left_join(sp.FOO, metadata, by = "SampleID")

# Remove low samples sites and unnecessary columns
sp_terr <- sp_metadata %>% 
  filter(!Territory %in% drop.sites) %>% 
  select(-c(Category, Individual))
  
# Split into list of dataframes for each territory
sp_terr_list <- split(sp_terr, sp_terr$Territory)

# Transpose dataframes, remove Territory row, and make species row names
sp_territories <- lapply(sp_terr_list, function(x) {
    data.table::transpose(x, make.names = "SampleID", keep.names = "Species") %>% 
    filter(Species != "Territory") %>% 
    remove_rownames %>% column_to_rownames(var="Species") %>% 
    mutate_if(is.character, as.numeric)})
```

## Individuals

```{r}
# Remove keep only individuals with IDs and samples >=8
sp_ind <- sp_metadata %>% 
  filter(Category == "SFCoy") %>% 
  select(-c(Category, Territory))

# Create character vector of individuals to retain that have at least 8 scats
individuals <- sp_ind %>% 
  group_by(Individual) %>% 
  tally() %>% 
  filter(n>=8) %>% 
  select(Individual)

coys.to.keep <- pull(individuals, Individual)

# Remove low sample individuals (n <8)
sp_ind_filt <- sp_ind %>% 
  filter(Individual %in% coys.to.keep)

# Split into list of dataframes for each individual
sp_ind_list <- split(sp_ind_filt, sp_ind_filt$Individual)

# Transpose dataframes, remove Territory row, and make species row names
sp_individuals <- lapply(sp_ind_list, function(x) {
    data.table::transpose(x, make.names = "SampleID", keep.names = "Species") %>% 
    filter(Species != "Individual") %>% 
    remove_rownames %>% column_to_rownames(var="Species") %>% 
    mutate_if(is.character, as.numeric)})
```

# Run iNEXT function

Calculate sample size, observed species richness, sample coverage, and diversity metrics for each territory and frequently sampled individual. Generate rarefaction curve plots.

## Territories

```{r}
# Colors
colors <- MetBrewer::met.brewer("Signac", 11)

# Calculate all diversity metrics
out_terr <- iNEXT(sp_territories, q=c(0,1,2), datatype = "incidence_raw", endpoint=200)

out_terr$DataInfo # get sample size (T), total number of incidences (U), observed species richness (S.obs), and sample coverage (SC) for each territory

# Shannon diversity
out_terr_shan <- iNEXT(sp_territories, q=1, datatype = "incidence_raw", endpoint=200)

# Plot Shannon rarefaction curves
df <- fortify(out_terr_shan, type=1)
df.point <- df[which(df$Method=="Observed"),]
df.line <- df[which(df$Method!="Observed"),]
df.line$Method <- factor(df.line$Method,
c("Rarefaction", "Extrapolation"),
c("Rarefaction", "Extrapolation"))

p_terr_shan_all <- ggplot(df, aes(x=x, y=y, colour=Assemblage)) +
geom_point(size=2, data=df.point) +
geom_line(aes(linetype=Method), lwd=0.8, data=df.line)+
ylab("Hill-Shannon Diversity")+
xlab("Number of fecal samples")+
theme_light()+
scale_colour_manual(values=colors)

p_terr_shan_all

# Simpson's diversity
out_terr_simp <- iNEXT(sp_territories, q=2, datatype = "incidence_raw", endpoint = 200)

# Plot Simpson rarefaction curves
df <- fortify(out_terr_simp, type=1)
df.point <- df[which(df$Method=="Observed"),]
df.line <- df[which(df$Method!="Observed"),]
df.line$Method <- factor(df.line$Method,
c("Rarefaction", "Extrapolation"),
c("Rarefaction", "Extrapolation"))

p_terr_simp_all <- ggplot(df, aes(x=x, y=y, colour=Assemblage)) +
geom_point(size=2, data=df.point) +
geom_line(aes(linetype=Method), lwd=0.8, data=df.line)+
ylab("Hill-Simpson Diversity")+
xlab("Number of fecal samples")+
theme_light()+
scale_colour_manual(values=colors)

p_terr_simp_all
```

## Individuals

```{r}
# Colors
colors <- as.vector(polychrome(22))

# Calculating sample coverage for each individual
estimateD(sp_individuals, datatype="incidence_raw", base="size", level=8, q=0)

# Calculate all diversity metrics
out_ind <- iNEXT(sp_individuals, q=c(0,1,2), datatype = "incidence_raw", endpoint=200)

out_ind$DataInfo # get sample size (T), total number of incidences (U), observed species richness (S.obs), and sample coverage (SC) for each individual

# Shannon Diversity
out_ind_shan <- iNEXT(sp_individuals, q=1, datatype = "incidence_raw", endpoint=50)

# Plot Shannon rarefaction curves
df <- fortify(out_ind_shan, type=1)
df.point <- df[which(df$Method=="Observed"),]
df.line <- df[which(df$Method!="Observed"),]
df.line$Method <- factor(df.line$Method,
c("Rarefaction", "Extrapolation"),
c("Rarefaction", "Extrapolation"))

p_ind_shan_all <- ggplot(df, aes(x=x, y=y, colour=Assemblage)) +
geom_point(size=2, data=df.point, alpha=0.7) +
geom_line(aes(linetype=Method), lwd=0.8, data=df.line, alpha=0.7)+
ylab("Hill-Shannon Diversity")+
xlab("Number of fecal samples")+
theme_light()+
scale_colour_manual(values=colors)

p_ind_shan_all

# Simpson Diversity
out_ind_simp <- iNEXT(sp_individuals, q=2, datatype = "incidence_raw", endpoint=50)

# Plot Simpson rarefaction curves
df <- fortify(out_ind_simp, type=1)
df.point <- df[which(df$Method=="Observed"),]
df.line <- df[which(df$Method!="Observed"),]
df.line$Method <- factor(df.line$Method,
c("Rarefaction", "Extrapolation"),
c("Rarefaction", "Extrapolation"))

p_ind_simpson_all <- ggplot(df, aes(x=x, y=y, colour=Assemblage)) +
geom_point(size=2, data=df.point, alpha=0.7) +
geom_line(aes(linetype=Method), lwd=0.8, data=df.line, alpha=0.7)+
ylab("Hill-Simpson Diversity")+
xlab("Number of fecal samples")+
theme_light()+
scale_colour_manual(values=colors)

p_ind_simpson_all
```
