---
title: "FilteringReads"
author: 
date: "2024-05-15"
output: html_document
---
# Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(data.table)
library(janitor)
```

# Read in ASV tables and remove sequences flagged for removal

For both batches, sequences comprising <0.001% of the total reads have already been removed. 

Drop sequences flagged for removal due to low percent identity or query cover.

```{r}
# Batch 1
seqs1 <- read.csv("Filtering-and-QC/batch1_reads.csv")
seqs1 <- seqs1 %>% 
  filter(Flag != "x") %>%
  select(FinalName,"BLANK_MB2":ncol(seqs1))

# Batch 2
seqs2 <- read.csv("Filtering-and-QC/batch2_reads.csv")
seqs2 <- seqs2 %>% 
  filter(Flag != "x") %>% 
  select(FinalName,"NTC1MB12":ncol(seqs2))
```

# Aggregate by taxa

```{r}
seqs1_count <- aggregate(. ~ FinalName, seqs1, sum)
seqs2_count <- aggregate(. ~ FinalName, seqs2, sum)
```

# Separate controls from samples

```{r}
# Batch 1
batch1_ctrls <- seqs1_count %>% select(-starts_with("S"))
batch1_sams <- seqs1_count %>% select(1, starts_with("S"))

# Batch 1
batch2_ctrls <- seqs2_count %>% select(-c(starts_with("R"), starts_with("S")))
batch2_sams <- seqs2_count %>% select(1, starts_with("R"), starts_with("S"))
```

Controls data frames are used to determine filtering thresholds. In this study, we used our negative and positive controls to set a filtering threshold of 4,400 reads for batch 1 and 150 reads for batch 2. See supplementary material figure S2 in the manuscript for more information. 

# Removing non-diet items from samples

Remove species that represent non-diet items including: low-level background contaminants outside the study region, positive control species, and host reads (Canis).

```{r}
# Create list of non-diet items to remove
nondiet <- c("Antilocapra americana", "Canis", "Cervus elaphus", "Chaetodipus", "Dama dama", "Dasypterus xanthinus", "Felidae", "Homo sapiens", "Hypomesus nipponensis", "Macaca mulatta", "Mirounga", "Monarchidae", "Odocoileus", "Oreortyx pictus", "Phoebastria albatrus", "Puma concolor", "Reithrodontomys raviventris",	"Spilogale gracilis", "Tamiasciurus", "Urocyon cinereoargenteus", "Ursus americanus")

# Create new dataframe that only keeps diet reads
batch1_prey <- batch1_sams %>% 
  filter(!FinalName %in% nondiet )

batch2_prey <- batch2_sams %>% 
  filter(!FinalName %in% nondiet )
```

# Calculate total number of reads per sample

```{r}
# Transpose dataframes
t_batch1 <- data.table::transpose(batch1_prey, make.names = "FinalName", keep.names = "SampleID")
t_batch2 <- data.table::transpose(batch2_prey, make.names = "FinalName", keep.names = "SampleID")

# Add a column calculating the total reads in each sample
total_batch1 <- t_batch1 %>% 
  mutate(Total = rowSums(across(where(is.numeric)), na.rm=TRUE), .after = "SampleID")

total_batch2 <- t_batch2 %>% 
  mutate(Total = rowSums(across(where(is.numeric)), na.rm=TRUE), .after = "SampleID")
```

# Apply filtering thresholds

For batch 1, remove all samples with <4,400 reads. For batch 2, remove all samples with <150 reads.

```{r}
# Remove unreliable samples (low amplification success) based on thresholds indicated above
rm_low_batch1 <- total_batch1 %>% 
  filter(Total >= 4400)

rm_low_batch2 <- total_batch2 %>% 
  filter(Total >= 150)
```

# Convert to proportions

Convert raw number of sequence reads into proportions for each diet item. 

```{r}
prop_batch1 <- rm_low_batch1 %>%
  select(!Total) %>% # remove the Total column
  adorn_percentages("row")

prop_batch2 <- rm_low_batch2 %>%
  select(!Total) %>% # remove the Total column
  adorn_percentages("row")
```

# Apply within-scat filtering

Remove diet items that represent <0.5% of the reads within a scat. However, to avoid errors of omission, manually include rare and/or small-bodied prey that is present in low proportions. These diet items include: *Junco hyemalis*, *Scapanus latimanus*, Trochilidae, and *Turdus migratorius* in batch 1 and *Didelphis virginiana* and *Sorex spp.* in batch 2.

```{r}
# Define function to remove diet items representing <0.5% of reads within a scat
suppress.to.zero <- function(x) ifelse(x <0.005, 0, x)

# Apply function to batch 1
rm.rare_batch1 <- prop_batch1 %>% 
  mutate(across(!SampleID &!`Junco hyemalis` &!`Scapanus latimanus` 
                &!Trochilidae &!`Turdus migratorius`, suppress.to.zero))

# Apply function to batch 2
rm.rare_batch2 <- prop_batch2 %>% 
  mutate(across(!SampleID &!`Didelphis virginiana` &!Sorex , suppress.to.zero))
```

# Normalize

After applying the suppress.to.zero function, proportions within scats no longer sum to one. The code below fixes this.

```{r}
# Pull out the non-numeric columns from the data-frame prior to transformation
character_batch1 <- rm.rare_batch1 %>% select(SampleID)
character_batch2 <- rm.rare_batch2 %>% select(SampleID)

# Normalize numeric columns using the apply function
norm_batch1 <-t(apply(rm.rare_batch1[,2:ncol(rm.rare_batch1)], 1, function(x) x/sum(x)))

norm_batch2 <-t(apply(rm.rare_batch2[,2:ncol(rm.rare_batch2)], 1, function(x) x/sum(x)))
```

# Create species-level RRA and FOO data sets for each batch

To convert from RRA to FOO, use the function defined below, which codes any amount of reads >0 as "1".

```{r}
# Add character columns back to the data frame to create the RRA data set
batch1_RRA <- data.frame(character_batch1, norm_batch1)
batch2_RRA <- data.frame(character_batch2, norm_batch2)

# Create function for converting RRA to FOO
convert.to.FOO <- function(x) ifelse(x >0, 1, 0)

# For all columns that are numeric, apply the function above
batch1_FOO <- batch1_RRA %>% mutate_if(is.numeric, convert.to.FOO)
batch2_FOO <- batch2_RRA %>% mutate_if(is.numeric, convert.to.FOO)
```

# Create functional group-level RRA and FOO data sets for each batch

```{r}
# Load functional group data
func.groups <- read.csv("Filtering-and-QC/Species_FunctionalGroups.csv")

# Pivot diet data and add functional group column
batch1_RRA_pivot <- batch1_RRA %>%
  pivot_longer(-SampleID, names_to = "Species", values_to = "RRA") %>% 
  left_join(func.groups, by = "Species")

batch2_RRA_pivot <- batch2_RRA %>%
  pivot_longer(-SampleID, names_to = "Species", values_to = "RRA") %>% 
  left_join(func.groups, by = "Species")

# Summarize the data by functional group and SampleID
batch1_RRA_func <- batch1_RRA_pivot %>%
  group_by(Group, SampleID) %>%
  summarise(Sum = sum(RRA, na.rm = TRUE)) %>% 
  pivot_wider(names_from = Group, values_from = Sum)

batch2_RRA_func <- batch2_RRA_pivot %>%
  group_by(Group, SampleID) %>%
  summarise(Sum = sum(RRA, na.rm = TRUE)) %>% 
  pivot_wider(names_from = Group, values_from = Sum)

# Convert RRA to FOO
batch1_FOO_func <- batch1_RRA_func %>% mutate_if(is.numeric, convert.to.FOO)
batch2_FOO_func <- batch2_RRA_func %>% mutate_if(is.numeric, convert.to.FOO)
```

# Combine batch 1 and batch 2 data sets into a single RRA and FOO data frames

After converting to RRA and FOO, data from each batch can be combined into a single data frame.

Species-level data:
```{r}
# Combine species RRA
sp.RRA_combined <- plyr::rbind.fill(batch1_RRA, batch2_RRA) %>% 
  replace(is.na(.), 0) # replace NAs with zeros
  
# Identify columns where all values are zeros
zero_columns <- sapply(sp.RRA_combined[, -1], function(col) all(col == 0))

# Remove columns where all values are zeros
sp.RRA <- sp.RRA_combined[, c(TRUE, !zero_columns)]

# Reorder alphabetically
new.order <- sort(colnames(sp.RRA[,2:ncol(sp.RRA)])) # generate alphabetical list of diet items
sp.RRA <- sp.RRA[,c("SampleID", new.order)] #reorder species columns to be alphabetical

# Convert RRA to FOO 
sp.FOO <- sp.RRA %>% mutate_if(is.numeric, convert.to.FOO)
```

Functional group-level data:
```{r}
# Combine functional group RRA
func.RRA <- plyr::rbind.fill(batch1_RRA_func, batch2_RRA_func) %>% 
  replace(is.na(.), 0) # replace NAs with zeros

# Convert RRA to FOO 
func.FOO <- func.RRA %>% mutate_if(is.numeric, convert.to.FOO)
```

# Final cleaning of diet data

Generate final data set by:
1.  Removing any samples with a known non-coyote origin (e.g., Cytb, genotyping, or intentionally collected dog samples)
2.  Removing samples that generated no *Canis* spp. reads
3.  Removing duplicate samples (PCR replication)

The final data set will contain samples only from SF coyotes and have no duplicates introduced by technical repliation.

Load metadata:
```{r}
# Load metadata
metadata <- read.csv("Filtering-and-QC/All_Metadata.csv")
  
# Load no Canis sample list
no.Canis <- read.csv("Filtering-and-QC/NoCanis_Scats.csv")
zero.read.scats <- no.Canis$SampleID # make character vector of samples to exclude

# Create duplicate sample removal list
dups.to.remove <- c("S20_3767","S20_3824","S20_3825","S20_3835","S20_3838","S21_0019","S21_0069","S21_0088","S21_0101","S21_0104")
```

Clean diet data:
```{r}
# Define input data frames
input <- list(func.RRA, func.FOO, sp.RRA, sp.FOO)

# Define function to clean data
clean.data <- function(x) {left_join(x, metadata, by = "SampleID") %>%
    filter((Cytb_Sp != "dog" & Cytb_Sp != "raccoon") %>% replace_na(T)) %>%
    filter((Geno.Sp != "dog") %>% replace_na(T)) %>%
    filter(!SampleID %in% zero.read.scats) %>%  # remove no Canis samples 
    filter(!SampleID %in% dups.to.remove)} # remove duplicates
    
# Run the function on all four input data frames
results <- lapply(input, clean.data)

# Save the output as four separate data frames
func.RRA.clean <- results[[1]]
func.FOO.clean <- results[[2]]
sp.RRA.output <- results[[3]]
sp.FOO.output <- results[[4]]

# Retain samples only
func.RRA.clean <- func.RRA.clean %>% select(-c("Replicate":"Cytb_Sp"))
func.FOO.clean <- func.FOO.clean %>% select(-c("Replicate":"Cytb_Sp"))
sp.RRA.output <- sp.RRA.output %>% select(-c("Replicate":"Cytb_Sp"))
sp.FOO.output <- sp.FOO.output %>% select(-c("Replicate":"Cytb_Sp"))

# Identify columns from species data frame where all values are zeros
zero_columns <- sapply(sp.RRA.output[, -1], function(col) all(col == 0))

# Remove columns where all values are zeros
sp.RRA.clean <- sp.RRA.output[, c(TRUE, !zero_columns)]
sp.FOO.clean <- sp.FOO.output[, c(TRUE, !zero_columns)]
```

# Save filtered and cleaned diet data for subsequent analyses

```{r}
write.csv(func.RRA.clean, "Filtering-and-QC/func.RRA.clean.csv")
write.csv(func.FOO.clean, "Filtering-and-QC/func.FOO.clean.csv")
write.csv(sp.RRA.clean, "Filtering-and-QC/sp.RRA.clean.csv")
write.csv(sp.FOO.clean, "Filtering-and-QC/sp.FOO.clean.csv")
```
