---
title: "PERMANOVA"
author: " "
date: "2024-05-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(vegan)
library(pairwiseAdonis)
library(BiodiversityR)
```

# Approach

To evaluate the significance of variation in vertebrate diet composition among territories, biological seasons, and individuals, we conducted a series of permutation-based multivariate analysis of variance tests (PERMANOVA; Anderson 2017) on the distance matrices. To test for seasonal differences, we used the date of sample collection to group scats according to biological season: mating: (November -- February), pupping (March -- June), and dispersal (July -- October). We first used all fecal samples to investigate differences in diet as a function of biological season and territory. Because not all scats yielded genotypes, and some samples came from the same individual, we verified that significant PERMANOVA results were not an artifact of pseudoreplication by conducting 1,000 trials in which we randomly sampled a single scat per individual.

Because we were interested in individual dietary variation, we investigated differences in diet at the individual level by comparing the most frequently sampled coyotes (≥8 scats). In this test, we included individual nested within family group as the main predictor and used a sequential (Type I) sum of squares.

We assessed differences in dispersion among all predictors to understand if dietary differences could be attributed to differences in group dispersion, a proxy for dietary niche breadth ("betadisper" and "permutest" in vegan). We ran all analyses with 999 permutations and conducted post-hoc tests for pairwise analysis with a Benjamini-Hochberg correction for multiple comparisons (via "pairwise.adonis" in the pairwiseAdonis package for composition and via "permutest" in the vegan package for dispersion.

# All samples

This dataset includes all samples assigned to a territory (n = 693) and therefore ignores genotypes.

## Prepare data

```{r}
# Load diet data
sp.RRA <- read.csv("Filtering-and-QC/sp.RRA.clean.csv") %>% select(-c(X))
sp.FOO <- read.csv("Filtering-and-QC/sp.FOO.clean.csv") %>% select(-c(X))

# Make sample list
sample_list <- sp.RRA %>% select(SampleID)

# Load metadata, add territories, biological seasons, and drop territories with low sample size
metadata <- read.csv("Filtering-and-QC/All_Metadata.csv")


# Load metadata, add territories, seasons, filter out transient territories and Sunset Reservoir, and select relevant columns
metadata <- read.csv("Filtering-and-QC/All_Metadata.csv") %>% 
  
  # Add territories 
  mutate(Territory = case_when(
    Site == "BERN" | Site == "STMA" ~ "Bernal",
    Site == "GLEN" | Site == "LAGU" | Site == "TWIN" | Site == "MTDA" | Site == "GGHP" ~ "Glen",
    Site == "SFGC" | Site == "OLYM" | Site == "LAKE" | Site == "FTFU" ~ "Merced",
    Site == "BVP" | Site == "CHP" ~ "Corona",
    Site == "TGPH" | Site == "HYDE" | Site == "FORT" ~ "Coit",
    Site == "GGPE" ~ "GGP - East",
    Site == "GGPW" | Site == "GGP" ~ "GGP - West",
    Site == "LEND" | Site == "LAFY" ~ "Land's End",
    Site == "MCLA" ~ "McLaren",
    Site == "PRES" ~ "Presidio",
    Site == "STFA" ~ "St. Francis",
    Site == "SUNS" ~ "Sunset Res",
    Site == "STGV" ~ "Stern Grove")) %>% 
  
  # Add biological seasons
  mutate(Biol.Season = case_when(
    Month == 3 | Month == 4 | Month == 5 | Month == 6 ~ "Pupping",
    Month == 7 | Month == 8 | Month == 9 | Month == 10 ~ "Dispersal",
    Month == 11 | Month == 12 | Month == 1 | Month == 2 ~ "Mating")) %>% 
  
  # Remove territories with low sample size
  filter(Territory != "Sunset Res" & Territory != "Stern Grove") %>%
  
  # Select only relevant columns
  select(c(SampleID, Biol.Season, Territory, Category, Individual)) %>%
  
  # Only retain metadata for samples in the final diet data frame
  inner_join(sample_list, by = "SampleID")
```

## Calculate dissimilarity matrix

Turn data frame into matrix:

```{r}
# Filter out samples from diet matrix that are not in the metadata
sp.RRA.filt <- left_join(metadata, sp.RRA, by = "SampleID")

names.filt <- sp.RRA.filt %>% select(SampleID) # create list of sample names

sp.RRA <- sp.RRA.filt %>% 
  select(-c(SampleID:Individual)) %>% 
  as.matrix

sp.FOO.filt <- left_join(metadata, sp.FOO, by = "SampleID")

sp.FOO <- sp.FOO.filt %>% 
  select(-c(SampleID:Individual)) %>% 
  as.matrix

# Add rownames
rownames(sp.RRA) <- names.filt$SampleID
rownames(sp.FOO) <- names.filt$SampleID
```

Calculate dissimilarity matrices:

```{r}
dist.sp.RRA <- vegdist(sp.RRA, method="bray")
dist.sp.FOO <- vegdist(sp.FOO, method="jaccard")
```

## Build models and run adonis2

```{r}
set.seed(1995)
permutations <- 999


# RRA
mod.all.RRA <- adonis2(dist.sp.RRA ~  Biol.Season + Territory,
                   data=metadata, permutations = permutations,
                   method="bray", by = "margin")
mod.all.RRA

# FOO
mod.all.FOO <- adonis2(dist.sp.FOO ~  Biol.Season + Territory,
                   data=metadata, permutations = permutations,
                   method="bray", by = "margin")

mod.all.FOO
```

## Pairwise comparisons of group means

Using a Benjamini-Hochberg p-value correction and running pairwise comparisons with pairwise.adnois from the package pairwiseAdonis.

```{r}
# Biological season pairwise comparisons
all.season.pairwise.RRA <- pairwise.adonis(x = dist.sp.RRA, factors = metadata$Biol.Season, p.adjust.m = "BH", perm = permutations)

all.season.pariwise.FOO <- pairwise.adonis(x = dist.sp.FOO, factors = metadata$Biol.Season, p.adjust.m= "BH", perm = permutations)

all.season.pairwise.RRA # 3 out of 3 are significantly different
all.season.pariwise.FOO # 3 out of 3 are significantly different

# Territory pairwise comparisons
all.terr.pairwise.RRA <- as.data.frame(pairwise.adonis(x = dist.sp.RRA, factors = metadata$Territory, p.adjust.m = "BH", perm = permutations))

all.terr.pairwise.FOO <- as.data.frame(pairwise.adonis(x = dist.sp.FOO, factors = metadata$Territory, p.adjust.m = "BH", perm = permutations))

table(all.terr.pairwise.RRA$sig != "") # 33 out of 55 comparisons are significantly different
table(all.terr.pairwise.FOO$sig != "") # 41 out of 55 comparisons are significantly different
```

## Testing for dispersion

When there are significant differences in dispersion, this can lead to false conclusions that group means are also different, when the difference may actually be attributed heterogeneity in dispersion.

```{r}
# Biological season betadisper
disp.season.RRA <- betadisper(dist.sp.RRA, metadata$Biol.Season)
disp.season.FOO <- betadisper(dist.sp.FOO, metadata$Biol.Season)

permutest(disp.season.RRA, pairwise=F) # significantly diff dispersion
permutest(disp.season.FOO, pairwise=F) # significantly diff dispersion

# Biological season comparison of group dispersions
pairwise.disp.season <- permutest(disp.season.RRA, pairwise=T)
pairwise.disp.season <- pairwise.disp.season$pairwise$permuted # extract p-values
pairwise.disp.season.pvals <- p.adjust(pairwise.disp.season, method="BH")# apply BH correction
table(pairwise.disp.season.pvals <= 0.05) # 2 out of 3 comparisons have different dispersion

pairwise.disp.season.FOO <- permutest(disp.season.FOO, pairwise=T)
pairwise.disp.season.FOO <- pairwise.disp.season.FOO$pairwise$permuted
pairwise.disp.season.FOO.pvals <- p.adjust(pairwise.disp.season.FOO, method="BH")
table(pairwise.disp.season.FOO.pvals <= 0.05) # 2 out of 3 comparisons have different dispersion


# Territory betadisper
disp.terr.RRA <- betadisper(dist.sp.RRA, metadata$Territory)
disp.terr.FOO <- betadisper(dist.sp.FOO, metadata$Territory)

permutest(disp.terr.RRA, pairwise=F) # significantly diff dispersion
permutest(disp.terr.FOO, pairwise=F) # significantly diff dispersion

# Territory comparison of group dispersions
pairwise.disp.terr <- permutest(disp.terr.RRA, pairwise=T)
pairwise.disp.terr <- pairwise.disp.terr$pairwise$permuted
pairwise.disp.terr.pvals <- p.adjust(pairwise.disp.terr, method="BH")
table(pairwise.disp.terr.pvals <= 0.05) # 16/55 comparisons have different dispersion

pairwise.disp.terr.FOO <- permutest(disp.terr.FOO, pairwise=T)
pairwise.disp.terr.FOO <- pairwise.disp.terr.FOO$pairwise$permuted
pairwise.disp.terr.FOO.pvals <- p.adjust(pairwise.disp.terr.FOO, method="BH")
table(pairwise.disp.terr.FOO.pvals <= 0.05)# 4/55 comparisons have diff dispersion
```

# Genotyped samples

Because not all scats yielded genotypes, and some samples came from the same individual, we verified that significant PERMANOVA results were not an artifact of pseudoreplication by conducting 1,000 trials in which we randomly sampled a single scat per individual (n = 111 total individuals).

To do this, we select only samples that have genotypes, sample one observation per individual, and re-run the PERMANOVA models. We repeat this process 1,000 times and then calculate average R2 values and the proportion of p-values that are significant for each covariate.

## Prepare data

```{r}
# Load diet data
sp.RRA <- read.csv("Filtering-and-QC/sp.RRA.clean.csv") %>% select(-c(X))
sp.FOO <- read.csv("Filtering-and-QC/sp.FOO.clean.csv") %>% select(-c(X))

# Make sample list
sample_list <- sp.RRA %>% select(SampleID)

# Load metadata, add territories, seasons, filter out transient territories and Sunset Reservoir, and select relevant columns
metadata.geno <- read.csv("Filtering-and-QC/All_Metadata.csv") %>% 
  
  # Add territories 
  mutate(Territory = case_when(
    Site == "BERN" | Site == "STMA" ~ "Bernal",
    Site == "GLEN" | Site == "LAGU" | Site == "TWIN" | Site == "MTDA" | Site == "GGHP" ~ "Glen",
    Site == "SFGC" | Site == "OLYM" | Site == "LAKE" | Site == "FTFU" ~ "Merced",
    Site == "BVP" | Site == "CHP" ~ "Corona",
    Site == "TGPH" | Site == "HYDE" | Site == "FORT" ~ "Coit",
    Site == "GGPE" ~ "GGP - East",
    Site == "GGPW" | Site == "GGP" ~ "GGP - West",
    Site == "LEND" | Site == "LAFY" ~ "Land's End",
    Site == "MCLA" ~ "McLaren",
    Site == "PRES" ~ "Presidio",
    Site == "STFA" ~ "St. Francis",
    Site == "SUNS" ~ "Sunset Res",
    Site == "STGV" ~ "Stern Grove")) %>% 
  
  # Add biological seasons
  mutate(Biol.Season = case_when(
    Month == 3 | Month == 4 | Month == 5 | Month == 6 ~ "Pupping",
    Month == 7 | Month == 8 | Month == 9 | Month == 10 ~ "Dispersal",
    Month == 11 | Month == 12 | Month == 1 | Month == 2 ~ "Mating")) %>% 
  
  # Remove territories with low sample size
  filter(Territory != "Sunset Res" & Territory != "Stern Grove") %>%
  
  # Keep only individuals with genotypes
  filter(Category == "SFCoy") %>% 
  
  # Select only relevant columns
  select(c(SampleID, Biol.Season, Territory, Individual)) %>%
  
  # Only retain metadata for samples in the final diet data frame
  inner_join(sample_list, by = "SampleID") %>% 
  
  # Make covarariates factors
  mutate(Biol.Season = as.factor(Biol.Season),
         Territory = as.factor(Territory),
         Individual = as.factor(Individual))
```

## Write function to sample data frame, prepare matrix, and run adonis2

Sample the genotype data frame such that each individual has one observation, generate a disimilarity matrix and run the model. Repeat the function as specified by the number of iterations, in this case 1,000. Then, write the results of all the models into a list. Pull out the p-values for each covariate, apply a p-value adjustment, and consider the results. I might want to include an excel table with all the tests as a supplementary file.

### RRA

Subsample down to one observation per individual and run adonis2 1000 times:

```{r}
# Set seed
set.seed(4)

# Set the number of iterations and permutations
num_iterations <- 1000
permutations = 999

# Define a list of model formulas
formulas <- list(
  formula1 = dist.RRA.geno ~ Biol.Season + Territory)

# Create a vector of by values corresponding to the formulas
by_values <- c("margin")

# Create an empty list to store model results and dissimilarity matrices
results_list <- vector("list", length = num_iterations)

# Perform the iterations
for (i in 1:num_iterations) {
  
  # Sample 1 observation per individual
  sampled_data <- metadata.geno %>%
    group_by(Individual) %>%
    sample_n(size = 1, replace = TRUE) %>%
    ungroup()

# Prepare and calculate dissimilarity matrix
  geno.RRA.sampled <- sampled_data %>% left_join(sp.RRA, by = "SampleID")

  names.geno.sampled <- geno.RRA.sampled %>% select(SampleID)
  
  sp.RRA.geno <- geno.RRA.sampled %>% 
  select(-c(SampleID:Year)) %>% 
  as.matrix
  
  # Add rownames
  rownames(sp.RRA.geno) <- names.geno.sampled$SampleID
  
  # Calculate dissimilarity
  dist.RRA.geno <- vegdist(sp.RRA.geno, method="bray")
  
# Run adonis2 on the sampled data for each model formula
  model_results <- lapply(1:length(formulas), function(j) {
    adonis_result <- adonis2(formulas[[j]], data = sampled_data, method="bray",
                             permutations=permutations, by = by_values[j])
    adonis_result
    })

# Store the model results in the list
  results_list[[i]] <- model_results
}
```

The piece of code above takes a long time to run. The model output is saved and can be loaded to save time:

```{r}
# Load results list if not re-running code above
results_list <- readRDS("PERMANOVA/PERMANOVA_sample_RRA.rds")
```

Summarize results:

```{r}
# Create an empty data frame to store the results
result_df <- data.frame(
  Trial = numeric(),
  Covariate = character(),
  F = numeric(),
  R2 = numeric(),
  P_value = numeric(),
  stringsAsFactors = FALSE
)

# Iterate through the results_list
for (i in seq_along(results_list)) {
  # Extract relevant information from each model output
  adonis_result <- results_list[[i]][[1]]

  # Extract values from the model output
  for (row_index in 1:2) {
  trial <- i
  covariate <- rownames(adonis_result)[row_index]  # Assuming the first row corresponds to Biol.Season
  F_value <- adonis_result$F[row_index]
  R2_value <- adonis_result$R2[row_index]
  p_value <- adonis_result$"Pr(>F)"[row_index]

  # Create a dataframe with the extracted values
  result_df <- rbind(result_df, c(trial, covariate, F_value, R2_value,
                                  p_value))
  # Rename columns in result_df
  colnames(result_df) <- c("Trial", "Covariate", "F", "R2", "Pr(>F)")
  
  # Change chr to numeric
  result_df <- result_df %>%
     mutate_at(vars(F, R2, `Pr(>F)`), as.numeric)
  }
}

# Save result to table (optional)
#write.csv(result_df, "PERMANOVA/PERMANOVA_sample_RRA.csv")

# Calculate average R2 value and proportion of p-values <= 0.05 for each covariate
summary_RRA <- result_df %>%
  group_by(Covariate) %>% 
  summarize(
    Mean_R2 = mean(R2),
    Count_Total = n(),
    Count_Sig_0.05 = sum(`Pr(>F)` <= 0.05),
    Count_Sig_0.1 = sum(`Pr(>F)` <= 0.1),
    Prop_Sig_0.05 = Count_Sig_0.05 / Count_Total,
    Prop_Sig_0.1 = Count_Sig_0.1 / Count_Total
  )

summary_RRA
```

### FOO

Subsample down to one observation per individual and run adonis2 1000 times:

```{r}
# Set seed
set.seed(4)

# Set the number of iterations and permutations
num_iterations <- 1000
permutations = 999

# Define a list of different formulas for your four models
formulas_FOO <- list(
  formula1 = dist.FOO.geno ~ Biol.Season + Territory)

# Create a vector of by values corresponding to the formulas
by_values <- c("margin")

# Create an empty list to store model results and dissimilarity matrices
results_list_FOO <- vector("list", length = num_iterations)

# Perform the iterations
for (i in 1:num_iterations) {
  
  # Sample 1 observation per individual
  sampled_data <- metadata.geno %>%
    group_by(Individual) %>%
    sample_n(size = 1, replace = TRUE) %>%
    ungroup()

# Prepare and calculate dissimilarity matrix
  geno.FOO.sampled <- sampled_data %>% left_join(sp.FOO, by = "SampleID")

  names.geno.FOO.sampled <- geno.FOO.sampled %>% select(SampleID)
  
  sp.FOO.geno <- geno.FOO.sampled %>% 
  select(-c(SampleID:Year)) %>% 
  as.matrix
  
  # Add rownames
  rownames(sp.FOO.geno) <- names.geno.FOO.sampled$SampleID
  
  # Calculate dissimilarity
  dist.FOO.geno <- vegdist(sp.FOO.geno, method="jaccard")
  
# Run adonis2 on the sampled data for each model formula
  model_results_FOO <- lapply(1:length(formulas_FOO), function(j) {
    adonis_result_FOO <- adonis2(formulas_FOO[[j]], data = sampled_data, method="jaccard", permutations=permutations, by = by_values[j])
    adonis_result_FOO
    })

# Store the model results in the list
  results_list_FOO[[i]] <- model_results_FOO
}
```

The piece of code above takes a long time to run. The model output is saved and can be loaded to save time:

```{r}
# Load results list if not re-running code above
results_list_FOO <- readRDS("PERMANOVA/PERMANOVA_sample_FOO.rds")
```

Summarize results:

```{r}
# Create an empty dataframe to store the results
result_df_FOO <- data.frame(
  Trial = numeric(),
  Covariate = character(),
  F = numeric(),
  R2 = numeric(),
  P_value = numeric(),
  stringsAsFactors = FALSE
)

# Iterate through the results_list
for (i in seq_along(results_list_FOO)) {
  # Extract relevant information from each model output
  adonis_result_FOO <- results_list_FOO[[i]][[1]]

  # Extract values from the model output
  for (row_index in 1:2) {
  trial <- i
  covariate <- rownames(adonis_result_FOO)[row_index]
  F_value <- adonis_result_FOO$F[row_index]
  R2_value <- adonis_result_FOO$R2[row_index]
  p_value <- adonis_result_FOO$"Pr(>F)"[row_index]

  # Create a dataframe with the extracted values
  result_df_FOO <- rbind(result_df_FOO, c(trial, covariate, F_value, R2_value,
                                  p_value))
  
  # Rename columns in result_df
  colnames(result_df_FOO) <- c("Trial", "Covariate", "F", "R2", "Pr(>F)")
  
  # Change chr to numeric
  result_df_FOO <- result_df_FOO %>%
     mutate_at(vars(F, R2, `Pr(>F)`), as.numeric)
  }
}

# Save result to table (optional)
#write.csv(result_df_FOO, "PERMANOVA/PERMANOVA_sample_FOO.csv")

# Calculate average R2 value and proportion of p-values <= 0.05 for each covariate
summary_FOO <- result_df_FOO %>%
  group_by(Covariate) %>% 
  summarize(
    Mean_R2 = mean(R2),
    Count_Total = n(),
    Count_Sig_0.05 = sum(`Pr(>F)` <= 0.05),
    Count_Sig_0.1 = sum(`Pr(>F)` <= 0.1),
    Prop_Sig_0.05 = Count_Sig_0.05 / Count_Total,
    Prop_Sig_0.1 = Count_Sig_0.1 / Count_Total
  )

summary_FOO
```

# Individual-level samples

Because we were interested in individual dietary variation, we investigated differences in diet at the individual level by comparing the most frequently sampled coyotes (≥8 scats) with high sample coverage. In this test, we included individual nested within family group as the main predictor and used a sequential (Type I) sum of squares with the BiodiversityR package.

## Prepare data

```{r}
# Load diet data
sp.RRA <- read.csv("Filtering-and-QC/sp.RRA.clean.csv") %>% select(-c(X))
sp.FOO <- read.csv("Filtering-and-QC/sp.FOO.clean.csv") %>% select(-c(X))

# Make sample list column
sample_list <- sp.RRA %>% select(SampleID)

# Load individual family group assignments
ids <- read.csv("Diet-Plots/IDs.csv")

# Make metadata column for only coyotes with 8 or more scats
metadata.ind <- read.csv("Filtering-and-QC/All_Metadata.csv") %>%  
  
  # Remove non-genotyped coyotes
  filter(Category == "SFCoy") %>% 
  
  # select only relevant columns
  select(c(SampleID, Individual)) %>%
  
  # Only keep metadata for samples in the final diet data frame
  inner_join(sample_list, by = "SampleID") %>% 
  
  # Add counts per individual
  add_count(Individual) %>% 
  
  # Add family group assignments
  left_join(ids, by = "Individual") %>% 
  
  # Filter coyotes with fewer than 8 scats
  filter(n >= 8) %>% 
  
  # Make character columns into factors
  mutate_if(is.character, as.factor)
```

## Calculate dissimilarity

```{r}
# Filter out samples from diet matrix that are not in the metadata
sp.RRA.filt.ind <- left_join(metadata.ind, sp.RRA, by = "SampleID")
sp.FOO.filt.ind <- left_join(metadata.ind, sp.FOO, by = "SampleID")

names.filt.ind.RRA <- sp.RRA.filt.ind %>% select(SampleID)
names.filt.ind.FOO <- sp.FOO.filt.ind %>% select(SampleID)

sp.RRA.ind <- sp.RRA.filt.ind %>% 
  select(-c(SampleID:Family.Group)) %>% 
  as.matrix

sp.FOO.ind <- sp.FOO.filt.ind %>% 
  select(-c(SampleID:Family.Group)) %>% 
  as.matrix

# Add rownames
rownames(sp.RRA.ind) <- names.filt.ind.RRA$SampleID
rownames(sp.FOO.ind) <- names.filt.ind.FOO$SampleID

# Calculate dissimilarity
dist.sp.RRA.ind <- vegdist(sp.RRA.ind, method="bray")
dist.sp.FOO.ind <- vegdist(sp.FOO.ind, method="jaccard")
```

## Build models and run adonis2

The sequential analysis offered by the Type I approach is appropriate for fully nested hierarchical models, for which there exists a natural ordering of the terms. In this case individual ID is nested with family group.

```{r}
set.seed(666)
permutations <- 999

# RRA
mod.nested.RRA <- nested.npmanova(dist.sp.RRA.ind ~ Family.Group + Individual,
                                data=metadata.ind, permutations=permutations, 
                                method="bray")

mod.nested.RRA 

# Calculate R2 of family group
mod.nested.RRA$SumsofSquares[1] / sum(mod.nested.RRA$SumsofSquares)

# Calculate R2 for individual ID
mod.nested.RRA$SumsofSquares[2] / sum(mod.nested.RRA$SumsofSquares)

# FOO
mod.nested.FOO <- nested.npmanova(dist.sp.FOO.ind ~ Family.Group + Individual,
                        data=metadata.ind, permutations=permutations,
                        method="jaccard")

mod.nested.FOO

# Calculate R2 of family group
mod.nested.FOO$SumsofSquares[1] / sum(mod.nested.FOO$SumsofSquares)

# Calculate R2 for individual ID
mod.nested.FOO$SumsofSquares[2] / sum(mod.nested.FOO$SumsofSquares)
```

## Pairwise comparisons of group means

```{r}
# Individuals
set.seed(4)
ind.pvals.BH <- pairwise.adonis(x = dist.sp.RRA.ind, factors = metadata.ind$Individual, p.adjust.m = "BH", perm = permutations)

table(ind.pvals.BH$sig != "") #94/231 had sig p values with the BH correction

set.seed(2)
ind.pvals.BH.FOO <- pairwise.adonis(x = dist.sp.FOO.ind, factors = metadata.ind$Individual, p.adjust.m = "BH", perm = permutations)

table(ind.pvals.BH.FOO$sig != "") #118/231 had sig p values with the BH correction


# Family Groups
set.seed(1995)
ind.pvals.BH.terr <- pairwise.adonis(x = dist.sp.RRA.ind, factors = metadata.ind$Family.Group, p.adjust.m = "BH", perm = permutations)

table(ind.pvals.BH.terr$sig != "") # 19/36 had sig p values with the BH correction

set.seed(1995)
ind.pvals.BH.terr.FOO <- pairwise.adonis(x = dist.sp.FOO.ind, factors = metadata.ind$Family.Group, p.adjust.m = "BH", perm = permutations)

table(ind.pvals.BH.terr.FOO$sig != "") # 22/36 had sig p values with the BH
```

## Testing for dispersion

```{r}
# Individuals betadisper
disp.ind.RRA <- betadisper(dist.sp.RRA.ind, metadata.ind$Individual)
set.seed(4)
permutest(disp.ind.RRA, pairwise=F) # significantly diff dispersion

disp.ind.FOO <- betadisper(dist.sp.FOO.ind, metadata.ind$Individual)
set.seed(4)
permutest(disp.ind.FOO, pairwise=F) # not significantly diff dispersion for FOO

# Individuals comparison of group dispersions
disp.pairwise.ind <- permutest(disp.ind.RRA, pairwise=T)
disp.pairwise.ind <- disp.pairwise.ind$pairwise$permuted # extract p-vals
disp.pairwise.ind.pvals <- p.adjust(disp.pairwise.ind, method="BH") # apply BH correction
table(disp.pairwise.ind.pvals <= 0.05) # only 13 out of 231 comparisons had significantly different dispersion


# Family groups betadisper
disp.fam <- betadisper(dist.sp.RRA.ind, metadata.ind$Family.Group)
set.seed(1)
permutest(disp.fam, pairwise=F) # significantly diff dispersion

disp.fam.FOO <- betadisper(dist.sp.FOO.ind, metadata.ind$Family.Group)
set.seed(1)
permutest(disp.fam.FOO, pairwise=F) # significantly diff dispersion

# Family group comparison of group dispersions
set.seed(1)
disp.pairwise.fam <- permutest(disp.fam, pairwise=T) 
disp.pairwise.fam <- disp.pairwise.fam$pairwise$permuted
disp.pairwise.fam.pvals <- p.adjust(disp.pairwise.fam, method="BH") 
table(disp.pairwise.fam.pvals <= 0.05) # 6 out of 36 comparisons have significantly different dispersion

set.seed(10)
disp.pairwise.fam.FOO <- permutest(disp.fam.FOO, pairwise=T)
disp.pairwise.fam.FOO <- disp.pairwise.fam.FOO$pairwise$permuted
disp.pairwise.fam.FOO.pvals <- p.adjust(disp.pairwise.fam.FOO, method="BH")
table(disp.pairwise.fam.FOO.pvals <= 0.05) # 3 out of 36 comparisons have significantly different dispersion
```
