---
title: "nMDS"
author: " "
date: "2024-05-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(vegan)
library(patchwork)
```

# Approach

We constructed Bray-Curtis (for RRA) and Jaccard (for FOO) dissimilarity matrices comparing each pair of fecal samples and ordinated the values with non-metric multidimensional scaling (nMDS) to visualize dietary differences among biological seasons, territories, and individuals using the vegan package in R. We selected the minimum number of dimensions (k = 3) in which stress of the ordination was ≤ 0.1.

# Biological seasons and territories

## Data preparation

```{r}
# Load diet data
sp.RRA <- read.csv("Filtering-and-QC/sp.RRA.clean.csv") %>% select(-c(X))
sp.FOO <- read.csv("Filtering-and-QC/sp.FOO.clean.csv") %>% select(-c(X))

# Load metadata
metadata <- read.csv("Filtering-and-QC/All_Metadata.csv")

# Prepare sample_lookup containing only information pertaining to samples included in diet analyses
sample_lookup <- left_join(sp.RRA, metadata, by = "SampleID") %>% 
  
  # Define seasons
  mutate(Biol.Season = case_when(
    Month == 3 | Month == 4 | Month == 5 | Month == 6 ~ "Pupping",
    Month == 7 | Month == 8 | Month == 9 | Month == 10 ~ "Dispersal",
    Month == 11 | Month == 12 | Month == 1 | Month == 2 ~ "Mating"
  )) %>% 
  
  # Define territories
  mutate(Territory = case_when(
    Site == "BERN" | Site == "STMA" ~ "Bernal",
    Site == "GLEN" | Site == "LAGU" | Site == "TWIN" | Site == "MTDA" | Site == "GGHP" ~ "Glen",
    Site == "SFGC" | Site == "OLYM" | Site == "LAKE" | Site == "FTFU" ~ "Merced",
    Site == "BVP" | Site == "CHP" ~ "Corona",
    Site == "TGPH" | Site == "HYDE" | Site == "FORT" ~ "Coit",
    Site == "GGPE" ~ "GGP - East",
    Site == "GGPW" | Site == "GGP" ~ "GGP - West",
    Site == "LEND" | Site == "LAFY" ~ "Land's End",
    Site == "MCLA" ~ "McLaren",
    Site == "PRES" ~ "Presidio",
    Site == "STFA" ~ "St. Francis",
    Site == "SUNS" ~ "Sunset Res",
    Site == "STGV" ~ "Stern Grove")) %>% 
  
  # Keep only relevant samples
  select(SampleID, Biol.Season, Territory, Individual, Genotype)
```

Remove seven outlier samples that prevent ordination from running: S22_0899, S21_0075, S21_0161, S20_3776, S21_1098, S21_0178, and S21_0023. Then, turn data frame into matrix.

```{r}
# Add rownames
rownames(sp.RRA) <- sp.RRA$SampleID
rownames(sp.FOO) <- sp.FOO$SampleID

# Define samples to drop
outliers <- c("S22_0899", "S21_0075", "S21_0161", "S20_3776", "S21_1098", "S21_0178", "S21_0023")

# Remove outliers and SampleID column and save as matrix
sp.RRA <- sp.RRA %>% 
  filter(!SampleID %in% outliers) %>% 
  select(-c(SampleID)) %>% 
  as.matrix()

sp.FOO <- sp.FOO %>% 
  filter(!SampleID %in% outliers) %>% 
  select(-c(SampleID)) %>% 
  as.matrix()
```

## Calculate dissimilarity matrices

```{r}
# Calculate dissimilarity
dist.sp.RRA <- vegdist(sp.RRA, method="bray")
dist.sp.FOO <- vegdist(sp.FOO, method="jaccard")

# Convert to matrix
dist.sp.RRA <- as.matrix(dist.sp.RRA, labels = T)
dist.sp.FOO <- as.matrix(dist.sp.FOO, labels = T)
```

## Run nMDS

Different features of the metaMDS command:

-   k = Number of dimensions.
-   trymax = How many NMDS runs should be performed in order to find a convergent solution (number of random starts). \*maxit = Indicates the maximum number of iterations each NMDS run can have in search of the lowest stress configuration.

Note, the metaMDS command takes a long time to run.

```{r}
set.seed(1)

# Run nMDS
nmds.sp.RRA <- metaMDS(dist.sp.RRA, k=3, autotransform = F, noshare = T, maxit = 999, trymax = 500, distance = "bray")

nmds.sp.FOO <- metaMDS(dist.sp.FOO, k=3, autotransform = F, noshare=T, maxit=999, trymax=500, distance="jaccard")

# Calculate stress score
nmds.sp.RRA$stress
nmds.sp.FOO$stress

# Produces Shepard's diagram to ensure that dissimilarities are well preserved in the reduced number of dimensions
stressplot(nmds.sp.RRA)
stressplot(nmds.sp.FOO)
```

Model output generated during data analysis was saved because running the metaMDS command takes a long time. These are saved as sp.RRA.rds and sp.FOO.rds. To save time, this model output can be loaded as follows:

```{r}
# Load metaMDS output
nmds.sp.RRA <- readRDS("nMDS/sp.RRA.rds")
nmds.sp.FOO <- readRDS("nMDS/sp.FOO.rds")
```

## Plot territory and season ordinations

Load data:

```{r}
# Load covariate data
covariates <- read.csv("Regression-Analyses/Territory_All_Covariates.csv")

# Load individual family group assignments
ids <- read.csv("Diet-Plots/IDs.csv")
```

Prep diet data for plotting:

```{r}
# Extract point positions
data.points.RRA <- as.data.frame(scores(nmds.sp.RRA))
data.points.FOO <- as.data.frame(scores(nmds.sp.FOO))

# Select low sample size territories to drop
sites.to.drop <- c("Stern Grove", "Sunset Res")

# Making plotting data frame with metadata associated
sp.RRA.plot <- data.points.RRA %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(sample_lookup,by = "SampleID") %>% 
  left_join(covariates, by="Territory") %>% 
  filter(!Territory %in% sites.to.drop)

sp.FOO.plot <- data.points.FOO %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(sample_lookup, by = "SampleID") %>% 
  left_join(covariates, by="Territory") %>% 
  filter(!Territory %in% sites.to.drop)

# Calculate group centroids for territories and seasons based on medians
centroid.sp.RRA <- sp.RRA.plot %>%
  group_by(Territory) %>% 
  mutate(centroid1 = median(NMDS1),
         centroid2 = median(NMDS2),
         centroid3 = median(NMDS3)) %>% 
  ungroup %>% 
  group_by(Biol.Season) %>% 
  mutate(centroid.season1 = median(NMDS1),
         centroid.season2 = median(NMDS2),
         centroid.season3 = median(NMDS3))

centroid.sp.FOO <- sp.FOO.plot %>% 
  group_by(Territory) %>% 
  mutate(centroid1 = median(NMDS1),
         centroid2 = median(NMDS2),
         centroid3 = median(NMDS3)) %>% 
  ungroup %>% 
  group_by(Biol.Season) %>% 
  mutate(centroid.season1 = median(NMDS1),
         centroid.season2 = median(NMDS2),
         centroid.season3 = median(NMDS3))
```

Make plots:

```{r}
# Define colors
colors <- c("#8DD3C7FF", "#FDDF11", "#BEBADAFF", "#FB8072FF", "#80B1D3FF", "#FDB462FF", "#B3DE69FF", "#FCCDE5FF", "#D9D9D9FF", "#BC80BDFF", "#CCEBC5FF", "black")

family_colors <- c("Bernal" = colors[1],
                   "Coit" = colors[2],
                   "Corona" = colors[3],
                   "GGP - East" = colors[4],
                   "GGP - West" = colors[5],
                   "Glen" = colors[6],
                   "Land's End" = colors[7],
                   "McLaren" = colors[8],
                   "Merced" = colors[9],
                   "Presidio" = colors[10],
                   "St. Francis" = colors[11],
                   "Transient" = colors[12])

## RRA

# Biological season plot
p.RRA.season <- sp.RRA.plot %>% 
  ggplot(aes(x=NMDS1, y=NMDS2, color=Biol.Season, fill=Biol.Season))+
  stat_ellipse(level=0.95, size=1, show.legend = FALSE)+
  geom_point(size=1.5, alpha=0.5, show.legend = FALSE)+
  geom_point(data=centroid.sp.RRA,
             mapping=aes(x=centroid.season1, y=centroid.season2),
                        color="black", shape=21, size=2, alpha=0.6,
                        show.legend=FALSE)+
  scale_color_manual(values=c("#66C2A5","#FC8D62","#8DA0CB"))+
  scale_fill_manual(values=c("#66C2A5", "#FC8D62", "#8DA0CB"))+
  theme_classic()+
  theme(plot.margin = margin(, 0.5, , , "cm"))+
  scale_x_continuous(breaks = c(-0.25, 0, 0.25),
                     labels = c("-0.25", "0", "0.25"))+
  scale_y_continuous(breaks = c(-0.25, 0, 0.25),
                     labels = c("-0.25", "0", "0.25"))+
  ggtitle("RRA")+
  theme(plot.title = element_text(hjust = 0.5))

# Territory plot
set.seed(1)
p.RRA.territory <- sp.RRA.plot %>% 
  ggplot(aes(x=NMDS1, y=NMDS2, color=Territory, fill=Territory))+
  stat_ellipse(level=.95, linewidth=1, show.legend = FALSE)+
  geom_point(size=1.5, alpha=0.5, show.legend = FALSE)+
  geom_jitter(data=centroid.sp.RRA,
             mapping=aes(x=centroid1, y=centroid2),
                        color="black", shape=21, size=2, alpha=0.6,
                        show.legend=FALSE, stat="unique", height=0.01)+
  scale_color_manual(values=family_colors)+
  scale_fill_manual(values=family_colors)+
  theme_classic()+
  theme(plot.margin = margin(, 0.5, , , "cm"))+
  scale_x_continuous(breaks = c(-0.25, 0, 0.25),
                     labels = c("-0.25", "0", "0.25"))+
  scale_y_continuous(breaks = c(-0.25, 0, 0.25),
                     labels = c("-0.25", "0", "0.25"))

## FOO

# Biological season plot
p.FOO.season <- sp.FOO.plot %>% 
  ggplot(aes(x=NMDS1, y=NMDS2, color=Biol.Season, fill=Biol.Season))+
  stat_ellipse(level=0.95, size=1, show.legend = FALSE)+
  geom_point(size=1.5, alpha=0.5, show.legend = FALSE)+
  geom_point(data=centroid.sp.FOO,
             mapping=aes(x=centroid.season1, y=centroid.season2),
                        color="black", shape=21, size=2, alpha=0.1, 
                        show.legend=FALSE)+
  scale_color_manual(values=c("#66C2A5","#FC8D62","#8DA0CB"))+
  scale_fill_manual(values=c("#66C2A5", "#FC8D62", "#8DA0CB"))+
  theme_classic()+
  theme(plot.margin = margin(, 0.5, , , "cm"))+
  scale_x_continuous(breaks = c(-0.25, 0, 0.25, 0.5),
                     labels = c("-0.25", "0", "0.25", "0.50"))+
  ggtitle("FOO")+
  theme(plot.title = element_text(hjust = 0.5))

# Territory plot
set.seed(1)
p.FOO.territory <- sp.FOO.plot %>% 
  ggplot(aes(x=NMDS1, y=NMDS2, color=Territory, fill=Territory))+
  stat_ellipse(level=0.95, linewidth=1, show.legend = FALSE)+
  geom_point(size=1.5, alpha=0.5, show.legend = FALSE)+
  geom_jitter(data=centroid.sp.FOO,
             mapping=aes(x=centroid1, y=centroid2),
                         color="black", shape=21, size=2,alpha=0.6, 
                        show.legend=FALSE, stat="unique", height=0.01)+
  scale_color_manual(values=family_colors)+
  scale_fill_manual(values=family_colors)+
  theme_classic()+
  theme(plot.margin = margin(, 0.5, , , "cm"))+
  scale_x_continuous(breaks = c(-0.25, 0, 0.25, 0.5),
                     labels = c("-0.25", "0", "0.25", "0.50"))
```

# Individuals

Pull out frequently samples individuals with eight or more scats and re-run metaMDS on individual-level data.

## Data preparation

```{r}
# Create vector of coyotes with 8 or more scats
individuals <- read.csv("Filtering-and-QC/sp.RRA.clean.csv") %>% select(-c(X)) %>% 
  left_join(metadata, by="SampleID") %>% 
  filter(Genotype=="Success") %>% 
  group_by(Individual) %>% 
  tally() %>% 
  filter(n>=8)
  
coys.to.keep <- as.vector(individuals$Individual)

# Load diet data and retain diet data only for individuals with 8 or more scats
sp.RRA.ind <- read.csv("Filtering-and-QC/sp.RRA.clean.csv") %>% select(-c(X)) %>% 
  left_join(metadata, by="SampleID") %>% 
  filter(Individual %in% coys.to.keep) %>% 
  select(SampleID:Zalophus.californianus)

sp.FOO.ind <- read.csv("Filtering-and-QC/sp.FOO.clean.csv") %>% select(-c(X)) %>% 
  left_join(metadata, by="SampleID") %>% 
  filter(Individual %in% coys.to.keep) %>% 
  select(SampleID:Zalophus.californianus)
```

Remove two outlier samples that prevent ordination from running: S22_0899 and S21_0023. Then, turn data frame into matrix.

```{r}
# # Add rownames
rownames(sp.RRA.ind) <- sp.RRA.ind$SampleID
rownames(sp.FOO.ind) <- sp.FOO.ind$SampleID

# Remove outliers and SampleID column and save as matrix
sp.RRA.ind <- sp.RRA.ind %>% 
  filter(SampleID != "S22_0899" & SampleID != "S21_0023") %>% 
  select(-c(SampleID)) %>% 
  as.matrix()

sp.FOO.ind <- sp.FOO.ind %>% 
  filter(SampleID != "S22_0899" & SampleID != "S21_0023") %>% 
  select(-c(SampleID)) %>% 
  as.matrix()
```

## Calculate dissimilarity

```{r}
# Calculate dissimilarity
dist.sp.RRA.ind <- vegdist(sp.RRA.ind, method="bray")
dist.sp.FOO.ind <- vegdist(sp.FOO.ind, method="jaccard")

# Convert to matrix
dist.sp.RRA.ind <- as.matrix(dist.sp.RRA.ind, labels = T)
dist.sp.FOO.ind <- as.matrix(dist.sp.FOO.ind, labels = T)
```

## Run nMDS

```{r}
set.seed(1)

# Run MDS
nmds.sp.RRA.ind <- metaMDS(dist.sp.RRA.ind, k=3, autotransform = F, noshare = T, maxit = 999, trymax = 250, distance = "bray")

nmds.sp.FOO.ind <- metaMDS(dist.sp.FOO.ind, k=3, autotransform = F, noshare = T, maxit = 999, trymax = 250, distance = "bray")

# Calculate stress score
nmds.sp.RRA.ind$stress
nmds.sp.FOO.ind$stress

# Produce Shepard's diagram to ensure that dissimilarities are well preserved in the reduced number of dimensions
stressplot(nmds.sp.RRA.ind)
stressplot(nmds.sp.FOO.ind)
```

Model output generated during data analysis was saved because running the metaMDS command takes a long time. These are saved as sp.RRA.ind.rds and sp.FOO.ind.rds. To save time, this model output can be loaded as follows:

```{r}
# Load metaMDS output
nmds.sp.RRA.ind <- readRDS("nMDS/sp.RRA.ind.rds")
nmds.sp.FOO.ind <- readRDS("nMDS/sp.FOO.ind.rds")
```

## Plot individual ordinations

Prep diet data for plotting:

```{r}
# Extract point positions
data.points.RRA.ind <- as.data.frame(scores(nmds.sp.RRA.ind))
data.points.FOO.ind <- as.data.frame(scores(nmds.sp.FOO.ind))

# Making plotting data frame with metadata associated
sp.RRA.ind.plot <- data.points.RRA.ind %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(sample_lookup,by = "SampleID") %>% 
  left_join(ids, by="Individual") %>% 
  mutate_at("Individual", as.factor)

sp.FOO.ind.plot <- data.points.FOO.ind %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(sample_lookup,by = "SampleID") %>% 
  left_join(ids, by="Individual") %>% 
  mutate_at("Individual", as.factor)

# Calculate group centroids for individuals based on medians
centroid.sp.RRA.ind <- sp.RRA.ind.plot %>%
  group_by(Individual) %>% 
  mutate(centroid1 = median(NMDS1),
         centroid2 = median(NMDS2),
         centroid3 = median(NMDS3))

centroid.sp.FOO.ind <- sp.FOO.ind.plot %>%
  group_by(Individual) %>% 
  mutate(centroid1 = median(NMDS1),
         centroid2 = median(NMDS2),
         centroid3 = median(NMDS3))
```

Make plots:

```{r}
# RRA
set.seed(1)
p.RRA.ind <- sp.RRA.ind.plot %>% 
  ggplot(aes(x=NMDS1, y=NMDS2, color=Family.Group, fill=Family.Group))+
  geom_point(size=1.5, alpha=0.5, show.legend = FALSE)+
  geom_jitter(data=centroid.sp.RRA.ind,
             mapping=aes(x=centroid1, y=centroid2),
                         color="black", shape=21, 
                        size=2, alpha=0.7, width=0.01, height=0.01, stat="unique",
             show.legend = FALSE)+
  scale_color_manual(values=family_colors)+  
  scale_fill_manual(values=family_colors)+
  theme_classic()+
  theme(plot.margin = margin(, 0.5, , , "cm"))+
  scale_x_continuous(breaks = c(-0.25, 0, 0.25, 0.5),
                     labels = c("-0.25", "0", "0.25", "0.5"),
                     limits = c(-0.35, 0.4))

# FOO
set.seed(3)
p.FOO.ind <- sp.FOO.ind.plot %>% 
  ggplot(aes(x=NMDS1, y=NMDS2, color=Family.Group, fill=Family.Group))+
  geom_point(size=1.5, alpha=0.5, show.legend = FALSE)+
  geom_jitter(data=centroid.sp.FOO.ind,
             mapping=aes(x=centroid1, y=centroid2),
                         color="black", shape=21, 
                        size=2, alpha=0.7, width=0.01, height=0.01, stat="unique",
             show.legend = FALSE)+
  scale_color_manual(values=family_colors)+  
  scale_fill_manual(values=family_colors)+
  theme_classic()+
  theme(plot.margin = margin(, 0.5, , , "cm"))
```

# Combine plots

```{r}
(p.RRA.season | p.FOO.season) /
(p.RRA.territory | p.FOO.territory)/
(p.RRA.ind | p.FOO.ind)
```
