---
title: "Diet-Plots"
author: 
date: "2024-05-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyr)
library(dplyr)
library(data.table)
library(plotrix)
library(ggpubr)
```

# Set Up

```{r}
# Define custom theme for plotting
theme_custom <- function() {
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        strip.background = element_rect(fill = "grey80", color = NA),
        axis.text.y = element_text(size=15),
        axis.text.x = element_text(size=15),
        axis.title.y = element_text(size=15),
        axis.title.x = element_text(size=15))}

# Define function for standard error of proportion
se.prop <- function(detections, total.scats) {
  p_hat <- detections / total.scats
  standard_error <- sqrt((p_hat * (1 - p_hat)) / total.scats)
  return(standard_error)
}

convert.to.FOO <- function(x) ifelse(x >0, 1, 0)
```

# Population-Level Diet

## Functional groups

Load data:

```{r}
# Load diet data
func.RRA <- read.csv("Filtering-and-QC/func.RRA.clean.csv") %>% select(-c(X))
func.FOO <- read.csv("Filtering-and-QC/func.FOO.clean.csv") %>% select(-c(X))

# Load metadata
metadata <- read.csv("Filtering-and-QC/All_Metadata.csv")
```

Calculate average RRA and FOO for each functional group:

```{r}
# Pivot dataframes
pivot.func.RRA <- func.RRA %>% 
  pivot_longer(-SampleID, names_to="Func.Group", values_to = "RRA")

pivot.func.FOO <- func.FOO %>% 
  pivot_longer(-SampleID, names_to="Func.Group", values_to = "FOO")

# Calculate average RRA and overall FOO
sum.RRA <- pivot.func.RRA %>%
  group_by(Func.Group) %>%
  summarize(meanRRA = 100*mean(RRA), se = 100*std.error(RRA), .groups="drop") 

sum.FOO <- pivot.func.FOO %>%
  group_by(Func.Group) %>%
  summarize(detections = sum(FOO), 
            total.scats=n(), 
            FOO=100*(detections/total.scats), 
            se.FOO = 100*se.prop(detections,total.scats))

# Combine data frames
all.dat <- full_join(sum.RRA, sum.FOO, by="Func.Group")
```

Plot:

```{r}
p.func.pop <- ggplot()+
  # RRA Data
  geom_col(aes(x=reorder(Func.Group, -meanRRA), y=meanRRA, fill=Func.Group), data=all.dat,
           alpha=0.5, width=0.3, position=position_nudge(x=0.1), show.legend = F)+
  geom_errorbar(aes(x=Func.Group, ymin=meanRRA-se, ymax=meanRRA+se), data=all.dat,
                width=.1, position=position_nudge(x=0.1))+
  # FOO Data
  geom_col(aes(x=reorder(Func.Group,-FOO), y=FOO, fill=Func.Group), data=all.dat, 
           width=0.3, position=position_nudge(x=-0.2), show.legend = F)+
  geom_errorbar(aes(x=Func.Group, ymin=FOO-se.FOO, ymax=FOO+se.FOO), data=all.dat,
                width=.1, position=position_nudge(x=-0.2))+
  # Graphics
  scale_fill_brewer(palette = "Set2")+
  scale_y_continuous(limits=c(0,103), expand = c(0, 0)) +
  ylab("Diet Item (FOO or RRA ± SE)")+
  theme_classic()+
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=10),
        axis.title.x = element_blank(),
        strip.background = element_blank(),
        strip.placement = "outside")+
  scale_x_discrete(labels = c("Small.Mammal" = "Small Mammals",
                              "Medium.Mammal" = "Medium-Sized Mammals",
                              "Bird" = "Birds",
                              "Herpetofauna" = "Herptiles",
                              "Marine.Aquatic" = "Marine Mammals"))

p.func.pop
```

## Species by Season

Load data:

```{r}
# Load diet data
sp.RRA <- read.csv("Filtering-and-QC/sp.RRA.clean.csv") %>% select(-c(X))
sp.FOO <- read.csv("Filtering-and-QC/sp.FOO.clean.csv") %>% select(-c(X))

# Load category data
species_categories <- read.csv("Diet-Plots/12S_species_categories.csv")
```

Generate data frames with average RRA and FOO per category:

```{r}
# Pivot main dataframe from wide to long format, add categories column, summarize by category, and pivot back
sp.RRA.categories <- sp.RRA %>%
  pivot_longer(-SampleID, names_to = "Species", values_to = "RRA") %>% 
  left_join(species_categories, by = "Species") %>% 
  group_by(Category, SampleID) %>%
  summarise(Sum = sum(RRA, na.rm = TRUE)) %>%
  pivot_wider(names_from = Category, values_from = Sum)

# Create FOO dataset
sp.FOO.categories <- sp.RRA.categories %>% mutate_if(is.numeric, convert.to.FOO)
```

Pull out the top 10 species from each data set. To do this, calculate a Total RRA or FOO column per category, sort, keeps the top 10 rows, and then pivot longer to create data frames that can be used for data visualization.

```{r}
# Pull out the top 10 categories for RRA and FOO:
top.RRA <- sp.RRA.categories %>% 
  data.table::transpose(make.names = "SampleID", keep.names = "Species") %>% 
  mutate(Total = rowSums(across(where(is.numeric)), na.rm=TRUE), .after = "Species") %>% 
  arrange(desc(Total)) %>%
  slice(1:10) %>% #keep top 10
  select(-c(Total)) %>%
  data.table::transpose(make.names = "Species", keep.names = "SampleID") %>%
  pivot_longer(-SampleID, names_to="Func.Category", values_to = "RRA")

top.FOO <- sp.FOO.categories %>% 
  data.table::transpose(make.names = "SampleID", keep.names = "Species") %>% 
  mutate(Total = rowSums(across(where(is.numeric)), na.rm=TRUE), .after = "Species") %>% 
  arrange(desc(Total)) %>%
  slice(1:10) %>% #keep top 10
  select(-c(Total)) %>%
  data.table::transpose(make.names = "Species", keep.names = "SampleID") %>%
  pivot_longer(-SampleID, names_to="Func.Category", values_to = "FOO")

# Define levels of top 10 species to maintain plotting order
RRA.levels <- rev(as.character(head(top.RRA$Func.Category, 10)))
FOO.levels <- (as.character(head(top.FOO$Func.Category, 10)))
```

To plot species by seasons, prepare season metadata:

```{r}
# Define seasons
season <- metadata %>% 
  mutate(Biol.Season = case_when(
    Month == 3 | Month == 4 | Month == 5 | Month == 6 ~ "Pupping",
    Month == 7 | Month == 8 | Month == 9 | Month == 10 ~ "Dispersal",
    Month == 11 | Month == 12 | Month == 1 | Month == 2 ~ "Mating"))  %>% 
  select(SampleID, Biol.Season) # pull out season data

# Join diet data to season data
top.RRA.season <- left_join(top.RRA, season, by = "SampleID")
top.FOO.season <- left_join(top.FOO, season, by = "SampleID")
```

Plot:

```{r}
# Calculate RRA and FOO per species per season
top.RRA.biol.season <- top.RRA.season %>%
  group_by(Biol.Season, Func.Category) %>%
  summarize(meanRRA = 100*mean(RRA), se = 100*std.error(RRA), .groups="drop") %>% 
  mutate(Func.Category = as.factor(Func.Category))

top.FOO.biol.season <- top.FOO.season %>% 
  group_by(Biol.Season, Func.Category) %>%  
  summarize(detections = sum(FOO), total.scats=n(), FOO=100*(detections/total.scats),
            se.FOO=100*se.prop(detections,total.scats))%>% 
  mutate(Func.Category = as.factor(Func.Category))

# Combine data frames
all.dat.biol.season <- full_join(top.RRA.biol.season, top.FOO.biol.season,
                                 by=c("Biol.Season", "Func.Category")) %>% 
  mutate(Func.Category = factor(Func.Category, levels = FOO.levels))


# Plot
p.biol.season <- ggplot(all.dat.biol.season) +
  
  # FOO Data
    geom_col(aes(x = Func.Category, y = FOO, fill = Biol.Season), 
             width = 0.4, position = position_dodge(width = 0.8)) +
    geom_errorbar(aes(x = Func.Category, ymin = FOO - se.FOO, ymax = FOO + se.FOO, 
                      fill=Biol.Season), 
                  width = 0.1, position = position_dodge(0.8))+
  # RRA Data
    geom_col(aes(x = as.numeric(Func.Category) + 0.135, y = meanRRA, 
                 fill = Biol.Season), alpha=0.5, width = 0.4, 
             position = position_dodge(width = 0.8))+
  geom_errorbar(aes(x=as.numeric(Func.Category) + 0.135, 
                    ymin = meanRRA-se, ymax = meanRRA+se, fill=Biol.Season), 
                width=0.1, position=position_dodge(0.8))+

  # Graphics
  scale_fill_brewer(palette = "Set2")+
  scale_y_continuous(limits=c(0,103), expand = c(0, 0))+
  ylab("")+ xlab(NULL)+
  theme_classic()+
  theme(axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.text.x = element_text(angle = 45, vjust = 1.05, hjust=1, size=10),
        strip.background = element_blank(),
        strip.placement = "outside")+
  scale_x_discrete(labels = c("Water.Birds" = "Water Birds", 
                              "Pigeons.Doves" = "Pigeons/Doves",
                              "Poultry" = "Chicken"))

p.biol.season
```

## Combined functional group and species by season plot

```{r}
ggarrange(p.func.pop, p.biol.season, 
          labels = c("(a)", "(b)"),
          label.x = c(0.15, 0.067),
          label.y = c(1, 1),
          ncol = 2, align="h", widths=c(0.3, 0.7))
```
