---
title: "Diet-Plots"
author: 
date: "2024-05-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(data.table)
library(plotrix)
library(ggpubr)
```

# Set Up

```{r}
# Define custom theme for plotting
theme_custom <- function() {
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        strip.background = element_rect(fill = "grey80", color = NA),
        axis.text.y = element_text(size=15),
        axis.text.x = element_text(size=15),
        axis.title.y = element_text(size=15),
        axis.title.x = element_text(size=15))}

# Define function for standard error of proportion
se.prop <- function(detections, total.scats) {
  p_hat <- detections / total.scats
  standard_error <- sqrt((p_hat * (1 - p_hat)) / total.scats)
  return(standard_error)
}

convert.to.FOO <- function(x) ifelse(x >0, 1, 0)
```

# Figure 3: Population-Level Bar Plots

## Functional groups

Load data:

```{r}
# Load diet data
func.RRA <- read.csv("Filtering-and-QC/func.RRA.clean.csv") %>% select(-c(X))
func.FOO <- read.csv("Filtering-and-QC/func.FOO.clean.csv") %>% select(-c(X))

# Load metadata
metadata <- read.csv("Filtering-and-QC/All_Metadata.csv")
```

Calculate average RRA and FOO for each functional group:

```{r}
# Pivot dataframes
pivot.func.RRA <- func.RRA %>% 
  pivot_longer(-SampleID, names_to="Func.Group", values_to = "RRA")

pivot.func.FOO <- func.FOO %>% 
  pivot_longer(-SampleID, names_to="Func.Group", values_to = "FOO")

# Calculate average RRA and overall FOO
sum.RRA <- pivot.func.RRA %>%
  group_by(Func.Group) %>%
  summarize(meanRRA = 100*mean(RRA), se = 100*std.error(RRA), .groups="drop") 

sum.FOO <- pivot.func.FOO %>%
  group_by(Func.Group) %>%
  summarize(detections = sum(FOO), 
            total.scats=n(), 
            FOO=100*(detections/total.scats), 
            se.FOO = 100*se.prop(detections,total.scats))

# Combine data frames
all.dat <- full_join(sum.RRA, sum.FOO, by="Func.Group")
```

Plot:

```{r}
p.func.pop <- ggplot()+
  # RRA Data
  geom_col(aes(x=reorder(Func.Group, -meanRRA), y=meanRRA, fill=Func.Group), data=all.dat,
           alpha=0.5, width=0.3, position=position_nudge(x=0.1), show.legend = F)+
  geom_errorbar(aes(x=Func.Group, ymin=meanRRA-se, ymax=meanRRA+se), data=all.dat,
                width=.1, position=position_nudge(x=0.1))+
  # FOO Data
  geom_col(aes(x=reorder(Func.Group,-FOO), y=FOO, fill=Func.Group), data=all.dat, 
           width=0.3, position=position_nudge(x=-0.2), show.legend = F)+
  geom_errorbar(aes(x=Func.Group, ymin=FOO-se.FOO, ymax=FOO+se.FOO), data=all.dat,
                width=.1, position=position_nudge(x=-0.2))+
  # Graphics
  scale_fill_brewer(palette = "Set2")+
  scale_y_continuous(limits=c(0,103), expand = c(0, 0)) +
  ylab("Diet Item (FOO or RRA ± SE)")+
  theme_classic()+
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=10),
        axis.title.x = element_blank(),
        strip.background = element_blank(),
        strip.placement = "outside")+
  scale_x_discrete(labels = c("Small.Mammal" = "Small Mammals",
                              "Medium.Mammal" = "Medium-Sized Mammals",
                              "Bird" = "Birds",
                              "Herpetofauna" = "Herptiles",
                              "Marine.Aquatic" = "Marine Mammals"))

p.func.pop
```

## Species by Season

Load data:

```{r}
# Load diet data
sp.RRA <- read.csv("Filtering-and-QC/sp.RRA.clean.csv") %>% select(-c(X))
sp.FOO <- read.csv("Filtering-and-QC/sp.FOO.clean.csv") %>% select(-c(X))

# Load category data
species_categories <- read.csv("Diet-Plots/12S_species_categories.csv")
```

Generate data frames with average RRA and FOO per category:

```{r}
# Pivot main dataframe from wide to long format, add categories column, summarize by category, and pivot back
sp.RRA.categories <- sp.RRA %>%
  pivot_longer(-SampleID, names_to = "Species", values_to = "RRA") %>% 
  left_join(species_categories, by = "Species") %>% 
  group_by(Category, SampleID) %>%
  summarise(Sum = sum(RRA, na.rm = TRUE)) %>%
  pivot_wider(names_from = Category, values_from = Sum)

# Create FOO dataset
sp.FOO.categories <- sp.RRA.categories %>% mutate_if(is.numeric, convert.to.FOO)
```

Pull out the top 10 species from each data set. To do this, calculate a Total RRA or FOO column per category, sort, keeps the top 10 rows, and then pivot longer to create data frames that can be used for data visualization.

```{r}
# Pull out the top 10 categories for RRA and FOO:
top.RRA <- sp.RRA.categories %>% 
  data.table::transpose(make.names = "SampleID", keep.names = "Species") %>% 
  mutate(Total = rowSums(across(where(is.numeric)), na.rm=TRUE), .after = "Species") %>% 
  arrange(desc(Total)) %>%
  slice(1:10) %>% #keep top 10
  select(-c(Total)) %>%
  data.table::transpose(make.names = "Species", keep.names = "SampleID") %>%
  pivot_longer(-SampleID, names_to="Func.Category", values_to = "RRA")

top.FOO <- sp.FOO.categories %>% 
  data.table::transpose(make.names = "SampleID", keep.names = "Species") %>% 
  mutate(Total = rowSums(across(where(is.numeric)), na.rm=TRUE), .after = "Species") %>% 
  arrange(desc(Total)) %>%
  slice(1:10) %>% #keep top 10
  select(-c(Total)) %>%
  data.table::transpose(make.names = "Species", keep.names = "SampleID") %>%
  pivot_longer(-SampleID, names_to="Func.Category", values_to = "FOO")

# Define levels of top 10 species to maintain plotting order
RRA.levels <- rev(as.character(head(top.RRA$Func.Category, 10)))
FOO.levels <- (as.character(head(top.FOO$Func.Category, 10)))
```

To plot species by seasons, prepare season metadata:

```{r}
# Define seasons
season <- metadata %>% 
  mutate(Biol.Season = case_when(
    Month == 3 | Month == 4 | Month == 5 | Month == 6 ~ "Pupping",
    Month == 7 | Month == 8 | Month == 9 | Month == 10 ~ "Dispersal",
    Month == 11 | Month == 12 | Month == 1 | Month == 2 ~ "Mating"))  %>% 
  select(SampleID, Biol.Season) # pull out season data

# Join diet data to season data
top.RRA.season <- left_join(top.RRA, season, by = "SampleID")
top.FOO.season <- left_join(top.FOO, season, by = "SampleID")
```

Plot:

```{r}
# Calculate RRA and FOO per species per season
top.RRA.biol.season <- top.RRA.season %>%
  group_by(Biol.Season, Func.Category) %>%
  summarize(meanRRA = 100*mean(RRA), se = 100*std.error(RRA), .groups="drop") %>% 
  mutate(Func.Category = as.factor(Func.Category))

top.FOO.biol.season <- top.FOO.season %>% 
  group_by(Biol.Season, Func.Category) %>%  
  summarize(detections = sum(FOO), total.scats=n(), FOO=100*(detections/total.scats),
            se.FOO=100*se.prop(detections,total.scats))%>% 
  mutate(Func.Category = as.factor(Func.Category))

# Combine data frames
all.dat.biol.season <- full_join(top.RRA.biol.season, top.FOO.biol.season,
                                 by=c("Biol.Season", "Func.Category")) %>% 
  mutate(Func.Category = factor(Func.Category, levels = FOO.levels))


# Plot
p.biol.season <- ggplot(all.dat.biol.season) +
  
  # FOO Data
    geom_col(aes(x = Func.Category, y = FOO, fill = Biol.Season), 
             width = 0.4, position = position_dodge(width = 0.8)) +
    geom_errorbar(aes(x = Func.Category, ymin = FOO - se.FOO, ymax = FOO + se.FOO, 
                      fill=Biol.Season), 
                  width = 0.1, position = position_dodge(0.8))+
  # RRA Data
    geom_col(aes(x = as.numeric(Func.Category) + 0.135, y = meanRRA, 
                 fill = Biol.Season), alpha=0.5, width = 0.4, 
             position = position_dodge(width = 0.8))+
  geom_errorbar(aes(x=as.numeric(Func.Category) + 0.135, 
                    ymin = meanRRA-se, ymax = meanRRA+se, fill=Biol.Season), 
                width=0.1, position=position_dodge(0.8))+

  # Graphics
  scale_fill_brewer(palette = "Set2")+
  scale_y_continuous(limits=c(0,103), expand = c(0, 0))+
  ylab("")+ xlab(NULL)+
  theme_classic()+
  theme(axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.text.x = element_text(angle = 45, vjust = 1.05, hjust=1, size=10),
        strip.background = element_blank(),
        strip.placement = "outside")+
  scale_x_discrete(labels = c("Water.Birds" = "Water Birds", 
                              "Pigeons.Doves" = "Pigeons/Doves",
                              "Poultry" = "Chicken"))

p.biol.season
```

## Combined functional group and species by season plot

```{r}
ggarrange(p.func.pop, p.biol.season, 
          labels = c("(a)", "(b)"),
          label.x = c(0.15, 0.067),
          label.y = c(1, 1),
          ncol = 2, align="h", widths=c(0.3, 0.7))
```

# Figure 4: Territory-Level Donut Plots

Prepare territory metadata:
```{r}
# Generate territory IDs from site data
territories <- metadata %>% 
  select(SampleID, Site) %>%  
  mutate(Territory = case_when(
    Site == "BERN" | Site == "STMA" ~ "Bernal",
    Site == "GLEN" | Site == "LAGU" | Site == "TWIN" | Site == "MTDA" | Site == 
      "GGHP" ~ "Glen",
    Site == "SFGC" | Site == "OLYM" | Site == "LAKE" | Site == "FTFU" ~ "Merced",
    Site == "BVP" | Site == "CHP" ~ "Corona",
    Site == "TGPH" | Site == "HYDE" | Site == "FORT" ~ "Coit",
    Site == "GGPE" ~ "GGP - East",
    Site == "GGPW" | Site == "GGP" ~ "GGP - West",
    Site == "LEND" | Site == "LAFY" ~ "Land's End",
    Site == "MCLA" ~ "McLaren",
    Site == "PRES" ~ "Presidio",
    Site == "STFA" ~ "St. Francis",
    Site == "SUNS" ~ "Sunset Res",
    Site == "STGV" ~ "Stern Grove"))

# Make character vector of sites to drop
drop.sites <- c("Stern Grove", "Sunset Res")

# Join diet data to territory data
func.RRA.terr <- left_join(pivot.func.RRA, territories, by = "SampleID")
```

The code below generates the donut plots for each territory based on RRA. These rings can be cropped and overlayed on a map of SF to produce Figure 4 in the manuscript.
```{r}
# Make donut plots
func.RRA.terr %>%
  filter(!Territory %in% drop.sites) %>% 
  group_by(Territory, Func.Group) %>%
  summarize(meanRRA = 100*mean(RRA), .groups="drop") %>% 
  ggplot(aes(x=2, y=meanRRA, fill=Func.Group)) +
  geom_col()+
  coord_polar(theta="y")+
  facet_wrap(~Territory, nrow=3, ncol=4)+
  scale_fill_brewer(palette = "Set2")+
  xlim(c(0.2, 2.5))+
  scale_y_continuous(breaks=NULL)+
  theme(legend.title=element_blank())+
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        legend.text = element_text(size=10),
        legend.key.size = unit(10, "pt"),
        strip.background = element_blank(),
        strip.text.x = element_text(size=10),
        panel.background = element_rect(fill = "white"),
        panel.grid = element_blank())+
  ylab("Average RRA (%)")+ xlab(" ")+
  labs(title = "RRA by Territory")
```

# Figure 6 - Individual-level bar plots organized into family groups

Load data:
```{r}
# Load diet data
func.RRA <- read.csv("Filtering-and-QC/func.RRA.clean.csv") %>% select(-c(X))

# Load metadata
metadata <- read.csv("Filtering-and-QC/All_Metadata.csv")

# Load individual family group assignments
ids <- read.csv("Diet-Plots/IDs.csv")
```

Merge the diet data to the metadata and filter out any scats that aren't assigned to an individual:
```{r}
# Merge diet data with metadata and keep only genotyped samples
func.RRA.ind <- left_join(func.RRA, metadata, by = "SampleID") %>% 
  filter(Category == "SFCoy")

# Create list of individuals to keep with >=8 or more scats
individuals <- func.RRA.ind %>% 
  group_by(Individual) %>% 
  tally() %>% 
  filter(n>=8)

coys.to.keep <- pull(individuals, Individual)

# Drop coyotes fewer than 8 scats and remove metadata
func.RRA.ind.filt <- func.RRA.ind %>% 
  filter(Individual %in% coys.to.keep) %>% 
  select(1:7, Individual)
```

Pivot data to longer format and calculate average RRA per individual:
```{r}
# Pivoting functional group data frames to longer format
pivot.func.RRA <- func.RRA.ind.filt %>% 
   pivot_longer(cols=c("Bird", "Small.Mammal", "Medium.Mammal", "Anthropogenic", "Herpetofauna", "Marine.Aquatic"), 
               names_to="Func.Group", values_to = "RRA")

# Calculate average RRA
func.RRA.ids <- pivot.func.RRA %>% 
  group_by(Individual, Func.Group) %>% 
  summarize(meanRRA = 100*mean(RRA), 
            se = 100*std.error(RRA), .groups="drop") %>% 
  left_join(ids, by = "Individual") 
```

Make plot:
```{r}
# Set names
names <- c("Anthropogenic", "Birds", "Herptiles", "Marine Mammals", "Medium-Sized Mammals", "Small Mammals")

# RRA
p.ind_stackedbar <- func.RRA.ids %>%
  ggplot(aes(x=reorder(Individual, Family.Group), y=meanRRA, fill=Func.Group)) +
  geom_col()+
  scale_y_continuous(expand = c(0,0))+
  facet_grid(~Family.Group, scales = "free", space = "free",
             labeller = labeller(Family.Group = label_wrap_gen(width = 1)))+
  scale_fill_brewer(palette = "Set2",
                    labels=names) +
  ylab("Diet Functional Group (RRA)") + xlab(NULL)+
  theme(axis.text.x = element_text(angle = 45, vjust = 1.15, hjust=1, size=8),
        axis.ticks.x = element_blank(),
        strip.text.x = element_text(size = 5, angle = 0, hjust=0.5, vjust=.5),
        strip.background.x = element_blank(),
        legend.title=element_blank(),
        legend.text = element_text(size = 8),  # Adjust the size of the legend text
        legend.key.size = unit(0.5, "cm"))

p.ind_stackedbar
```

