---
title: "Diet-Specialization"
author: " "
date: "2024-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RInSp)
library(mgcv)
library(ggrepel)
library(cowplot)

# Set custom plot theme
theme_custom <- function() {
  theme_classic() +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        strip.background = element_rect(fill = "grey80", color = NA),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        axis.title.y = element_text(size=12),
        axis.title.x = element_text(size=12))}
```

# Prepare data

Format data for analyses in RInSp.

```{r}
# Load diet data
sp.RRA <- read.csv("Filtering-and-QC/sp.RRA.clean.csv") %>% select(-c(X))

# Load covariate data
covs <- read.csv("Regression-Analyses/Territory_All_Covariates.csv")

# Load family group data
fams <- read.csv("Diet-Plots/IDs.csv")

# Make sample list
sample_list <- sp.RRA %>% select(SampleID)

# Load metadata, add territories, seasons, and select relevant columns
metadata <- read.csv("Filtering-and-QC/All_Metadata.csv") %>% 
  
  # Add territories 
  mutate(Territory = case_when(
    Site == "BERN" | Site == "STMA" ~ "Bernal",
    Site == "GLEN" | Site == "LAGU" | Site == "TWIN" | Site == "MTDA" | Site == "GGHP" ~ "Glen",
    Site == "SFGC" | Site == "OLYM" | Site == "LAKE" | Site == "FTFU" ~ "Merced",
    Site == "BVP" | Site == "CHP" ~ "Corona",
    Site == "TGPH" | Site == "HYDE" | Site == "FORT" ~ "Coit",
    Site == "GGPE" ~ "GGP - East",
    Site == "GGPW" | Site == "GGP" ~ "GGP - West",
    Site == "LEND" | Site == "LAFY" ~ "Land's End",
    Site == "MCLA" ~ "McLaren",
    Site == "PRES" ~ "Presidio",
    Site == "STFA" ~ "St. Francis",
    Site == "SUNS" ~ "Sunset Res",
    Site == "STGV" ~ "Stern Grove")) %>% 
  
  # Add biological seasons
  mutate(Biol.Season = case_when(
    Month == 3 | Month == 4 | Month == 5 | Month == 6 ~ "Pupping",
    Month == 7 | Month == 8 | Month == 9 | Month == 10 ~ "Dispersal",
    Month == 11 | Month == 12 | Month == 1 | Month == 2 ~ "Mating")) %>% 
  
  # Select only relevant columns
  select(c(SampleID, Biol.Season, Territory, Category, Individual)) %>%
  
  # Only retain metadata for samples in the final diet data frame
  inner_join(sample_list, by = "SampleID") %>% 
  
  # Turn character columns to factors
  mutate(Biol.Season = as.factor(Biol.Season),
         Territory = as.factor(Territory),
         Category = as.factor(Category),
         Individual = as.factor(Individual))
```

# Population BIC/TNW

Calculates the average relative read abundance per diet item per individual and transform these proportions to integers by multiplying by 1,000 and rounding to the nearest whole number. These integer data reflect the relative read abundances ascribed to each individual. Then, overall BIC/TNW is calculated for the population.

```{R}
# Create data frame of averaged RRA values per individual and convert to integers
geno_RRA <- inner_join(metadata, sp.RRA, by = "SampleID") %>% 
  filter(Category == "SFCoy") %>% # only include scats with individual IDs 
  group_by(Individual) %>% 
  summarize(across(where(is.numeric), ~ round(mean(.) * 1000)))  %>% # multiply by 1000 & round to nearest whole number
  mutate(ID = seq(1, nrow(.))) %>% 
  relocate(ID, .before = Individual) %>% 
  left_join(fams, by="Individual") %>% # add individual family group assignments
  relocate(Family.Group, .after=ID) %>% 
  relocate(Sex, .after = Family.Group) %>% 
  mutate(FamID = as.factor(Family.Group), .before=Family.Group) %>% 
  as.data.frame()

# Create RInSp object
rinsp_geno_RRA = import.RInSp(geno_RRA, row.names = 1, data.type = "integer", info.cols = c(2:5))

# WIC/TNW
WT_geno_RRA = WTdMC(rinsp_geno_RRA, replicates=999, print.ris=TRUE, pop.diet = "average")

# Extract metrics
pop_WIC.TNW <- WT_geno_RRA$WonT
pop_BIC.TNW <- 1-WT_geno_RRA$WonT
```

 Using Roughgarden's 1979 equations, based on Shannon-Weaver diversity index: 
 Within-individual component          =  0.9540732
 Between-individual component         =  1.053788
 Total Niche Width for the population =  2.007861
 The value of WIC/TNW is:  0.4751689
 The p-value is:  0.001 

# Territory-level BIC/TNW

Calculate BIC/TNW for each family group of coyotes.

```{r}
# Define a function to process each territory
process_familygroup <- function(family_name) {
  # Filter, group, summarize, and format data
  geno_data <- inner_join(metadata, sp.RRA, by = "SampleID") %>%
    left_join(fams, by="Individual") %>% 
    filter(Category == "SFCoy", Family.Group == family_name) %>%
    group_by(Individual) %>%
    summarize(across(where(is.numeric), ~ round(mean(.) * 10000))) %>%
    mutate(ID = seq(1, nrow(.))) %>%
    relocate(ID, .before = Individual) %>%
    as.data.frame()
  
  # Get count of number of individuals included per territory
  row_count <- nrow(geno_data)
  
  # Create RInSp object
  rinsp_obj <- import.RInSp(geno_data, row.names = 1, data.type = "integer", info.cols = 2)
  
  # Calculate WIC/TNW index
  WT_result <- WTdMC(rinsp_obj, replicates = 999, print.ris = TRUE, pop.diet = "average")
  
  # Return results for WonT and p-value
  list(Family.Group = family_name, WIC.TNW = WT_result$WonT, p_value = WT_result$p.value,
       Num.Ind = row_count)
}

# Get unique family group names
unique_fams <- unique(fams$Family.Group)

# Apply the function to each family group and store results
results <- map(unique_fams, process_familygroup)

# Calculate the number of scats used in each family group
territory_counts <- inner_join(metadata, sp.RRA, by = "SampleID") %>%
    left_join(fams, by="Individual") %>% 
    filter(Category == "SFCoy") %>% 
    filter(Family.Group != "Transient") %>% 
    group_by(Family.Group) %>% 
    mutate(Num.Scats = n()) %>% 
    relocate(Family.Group, .after = SampleID) %>% 
    relocate(Num.Scats, .after = Family.Group) %>% 
    distinct(Family.Group, Num.Scats)

# Convert list of results to a data frame and add covariate data
results_df <- bind_rows(results) %>% 
  mutate(BIC.TNW = 1-WIC.TNW, .after=WIC.TNW) %>% 
  filter(Family.Group != "Transient") %>% # drop transient individual
  left_join(covs, by = c("Family.Group" = "Territory")) %>%  # join environmental covariates
  left_join(territory_counts, by="Family.Group") %>% # join sample counts
  mutate(Fam.Names = case_when( # add custom family group names for plotting
    Family.Group == "Bernal" ~ "Bernal Hill",
    Family.Group == "Coit" ~ "Coit Tower",
    Family.Group == "Glen" ~ "Glen Canyon",
    Family.Group == "Corona" ~ "Corona Heights Park",
    Family.Group == "GGP - East" ~ "Golden Gate Park - East",
    Family.Group == "St. Francis" ~ "St. Francis Wood",
    Family.Group == "GGP - West" ~ "Golden Gate Park - West",
    Family.Group == "McLaren" ~ "McLaren Park",
    TRUE ~ Family.Group))

# View the results
arrange(results_df, BIC.TNW)
```

Model and plot results:
```{r}
# Model BIC/TNW as a function of ISA and number of scats
mod.BIC.ISA.beta.gam <- gam(BIC.TNW ~ ISA + Num.Scats, family=betar(link="logit"), data = results_df)
summary(mod.BIC.ISA.beta.gam)

# Generate new ISA and Num.Scats data for making predictions
new_data <- data.frame(ISA = seq(0, 100, by=0.1),
                       Num.Scats = round(mean(results_df$Num.Scats)))

# Make predictions on the link scale, calculate confidence interval boundaries on the link scale, then back transform using the inverse link to get boundaries on the response scale
gam_preds_ISA <- as.data.frame(mgcv::predict.gam(mod.BIC.ISA.beta.gam, newdata = new_data, se.fit = TRUE, type="link")) %>% 
  mutate(ISA = new_data$ISA, .before="fit") %>% 
  mutate(upr = fit + (1.96 * se.fit),
         lwr = fit - (1.96 * se.fit),
         BIC.TNW = mod.BIC.ISA.beta.gam$family$linkinv(fit),
         upr_adj = mod.BIC.ISA.beta.gam$family$linkinv(upr),
         lwr_adj = mod.BIC.ISA.beta.gam$family$linkinv(lwr))

# Plot BIC/TNW as a function of ISA
ggplot() +
  
  # Add regression line for predictions of BIC/TNW for each ISA value
  geom_smooth(data = gam_preds_ISA, aes(x = ISA, y = BIC.TNW,
                                   ymin = lwr_adj, ymax = upr_adj), stat = 'identity',
            color="#696464", fill="darkgrey", alpha=0.5, linewidth=0.5)+
  
  # Add population-level BIC/TNW
   geom_hline(yintercept=(pop_BIC.TNW), linetype = "dashed")+
  
  # Add BIC/TNW values for each family group
  geom_point(data = results_df, aes(x = ISA, y = BIC.TNW), size=0.5)+
  
  # Add family group labels
  geom_text_repel(data = results_df, aes(x = ISA, y = BIC.TNW, label = Fam.Names),
                  size = 2,    # Smaller label size
                  segment.color = "grey50", segment.size = 0.3,
                  direction="both") +
  
  # Set graphics
  theme_custom()+
  ylab("BIC/TNW")+xlab("Impervious Surface Cover")+
  scale_x_continuous(limits=c(0,101), expand=c(0, 0))+
  scale_y_continuous(limits=c(0,1))

# Save plot
#ggsave("RInSp/BIC.TNW.plot.png", dpi=600, height=3, width=4)
```

# Individual data

Calculate Proportional Similarity index (PSi) values for each coyote with eight or more scats.

```{r}
# Create subsampled dataset with 8 scats per individual and raw sequence reads
ind_samples <- inner_join(metadata, sp.RRA, by = "SampleID") %>% 
  filter(Category == "SFCoy") %>% 
  select(c(SampleID, Individual)) %>% 
  group_by(Individual) %>%  
  mutate(count = n(), .after="Individual") %>% 
  filter(count >=8) %>% 
  distinct(Individual, count)

# Calculate PSi values for every coyote in the data set
PSi_all = PSicalc(rinsp_geno_RRA, pop.diet = "average", exclude = FALSE, replicates=999, precision = 1e-9)

# Create list of 22 frequently sampled coyotes with 8 or more scats
coyote.names <- ind_samples$Individual

# Extract PSi values for 22 coyotes with 8 or more scats
PSi_all_df <- data.frame(PSi = PSi_all$PSi) %>% 
  mutate(ID = seq(1,nrow(.))) %>% 
  left_join(geno_RRA %>% select(c(ID, Individual, Family.Group)), by="ID") %>% 
  filter(Individual %in% coyote.names) %>% 
  left_join(ind_samples, by="Individual")

# View the results
arrange(PSi_all_df, PSi)

# Plot as function of sample size
ggplot(PSi_all_df)+
  geom_point(aes(x=count, y=PSi))+
  geom_smooth(aes(x=count, y=PSi),method="lm")+
  scale_x_continuous(breaks=c(seq(0,25,by=1)))+
  theme_custom()

# Test the effect of sample size on PSi values
mod.PSi<-  gam(PSi ~ count, family=betar(link="logit"), data = PSi_all_df)
summary(mod.PSi) # not significant
```

## Figure 4

Overlay PSi values onto stacked bar plot visualizing diet variation among frequently sampled coyotes. NOTE - prior to running this code the stacked bar plot (p.ind_stackedbar) from the Diet-Plots.Rmd file must be generated. This can be produced by running the code in Diet-Plot.Rmd from line 308.

```{r}
# Plot PSi values faceted by family group
p.PSi <- PSi_all_df %>%
  ggplot(aes(x=reorder(Individual, Family.Group), y=PSi)) +
  geom_point(size=1.5)+
  scale_y_continuous(expand = c(0,0), limits=c(0,1), position="right")+
  facet_grid(~Family.Group, scales = "free", space = "free",
             labeller = labeller(Family.Group = label_wrap_gen(width = 1)))+
  ylab("PSi") + xlab(NULL)+
  theme_void() +
  theme(strip.text.x = element_blank(),
        axis.title.y = element_text(angle=270, size=10),
        axis.text.y = element_text(margin = margin(r = 5), size=8, color="#504f4f"),
        axis.ticks.y = element_line(),
        axis.ticks.length.y = unit(0.08, "cm"),
        legend.title=element_blank())

# Move legend out of the way on stacked bar plot
p.diet <- p.ind_stackedbar +
  theme(legend.box.spacing = unit(1, "cm"),
        axis.text.y=element_text(color="#504f4f"),
        axis.text.x=element_text(color="#504f4f"))

# Align plots
p.aligned <- cowplot::align_plots(p.diet, p.PSi, align = c("hv"), axis = 'rlbt')

# Overlay plots
ggdraw() +
  draw_plot(p.aligned[[1]], 0, 0, 1, 1) +
  draw_plot(p.aligned[[2]], 0, 0, 1, 1)

# Save plot
#ggsave("diet.PSi.png", dpi=600, height=4, width=6.5)
```
