---
title: "Diet-Specialization"
author: "Tali Caspi"
date: "2024-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RInSp)
library(bayestestR)

# Define custom theme for plotting
theme_custom <- function() {
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        strip.background = element_rect(fill = "grey80", color = NA),
        axis.text.y = element_text(size=15),
        axis.text.x = element_text(size=15),
        axis.title.y = element_text(size=15),
        axis.title.x = element_text(size=15))}
```

# Prepare data

```{r}
# Load diet data
sp.RRA <- read.csv("Filtering-and-QC/sp.RRA.clean.csv") %>% select(-c(X))
sp.FOO <- read.csv("Filtering-and-QC/sp.FOO.clean.csv") %>% select(-c(X))
func.RRA <- read.csv("Filtering-and-QC/func.RRA.clean.csv") %>% select(-c(X))
rawseqs.clean <- read.csv("Filtering-and-QC/raw.seqs.clean.csv") %>% select(-c(X))

# Load covariate data
covs <- read.csv("Regression-Analyses/Territory_All_Covariates.csv")

# Load family group data
fams <- read.csv("Diet-Plots/IDs.csv")

# Make sample list
sample_list <- rawseqs.clean %>% select(SampleID)

# Load metadata, add territories, seasons, and select relevant columns
metadata <- read.csv("Filtering-and-QC/All_Metadata.csv") %>% 
  
  # Add territories 
  mutate(Territory = case_when(
    Site == "BERN" | Site == "STMA" ~ "Bernal",
    Site == "GLEN" | Site == "LAGU" | Site == "TWIN" | Site == "MTDA" | Site == "GGHP" ~ "Glen",
    Site == "SFGC" | Site == "OLYM" | Site == "LAKE" | Site == "FTFU" ~ "Merced",
    Site == "BVP" | Site == "CHP" ~ "Corona",
    Site == "TGPH" | Site == "HYDE" | Site == "FORT" ~ "Coit",
    Site == "GGPE" ~ "GGP - East",
    Site == "GGPW" | Site == "GGP" ~ "GGP - West",
    Site == "LEND" | Site == "LAFY" ~ "Land's End",
    Site == "MCLA" ~ "McLaren",
    Site == "PRES" ~ "Presidio",
    Site == "STFA" ~ "St. Francis",
    Site == "SUNS" ~ "Sunset Res",
    Site == "STGV" ~ "Stern Grove")) %>% 
  
  # Add biological seasons
  mutate(Biol.Season = case_when(
    Month == 3 | Month == 4 | Month == 5 | Month == 6 ~ "Pupping",
    Month == 7 | Month == 8 | Month == 9 | Month == 10 ~ "Dispersal",
    Month == 11 | Month == 12 | Month == 1 | Month == 2 ~ "Mating")) %>% 
  
  # Select only relevant columns
  select(c(SampleID, Biol.Season, Territory, Category, Individual)) %>%
  
  # Only retain metadata for samples in the final diet data frame
  inner_join(sample_list, by = "SampleID") %>% 
  
  # Turn character columns to factors
  mutate(Biol.Season = as.factor(Biol.Season),
         Territory = as.factor(Territory),
         Category = as.factor(Category),
         Individual = as.factor(Individual))
```

# All scats with genotypes (n=525)

## Subsampling approach using sequence reads and proportions

Using a data frame containing sequence reads per diet item (integer data), randomly subsample the dataset so that there is a single scat per individual. Then, calculate WIC/TNW and PSi values and associate p-values. Iterate this process 1,000 times and calculate an means specialization indices with associated confidence intervals as well as the proportion of p-values that are significant. Use pop.diet = "average" to ensure that individual diets are first converted into proportions, and only then are these proportions averaged for each resource.

```{r}
# Prepare sequence read data set 
geno_samples <- inner_join(metadata, rawseqs.clean, by="SampleID") %>% 
  filter(Category == "SFCoy")

# Randomly sample one scat per individual
sampled_data <- geno_samples %>% 
    group_by(Individual) %>%
    sample_n(size = 1, replace = TRUE) %>%
    ungroup() %>% 
   mutate(ID = seq(1,nrow(.))) %>% 
    relocate(ID, .before=SampleID) %>% 
    as.data.frame()

# Create RInSp object
rinsp_geno_subsamples = import.RInSp(sampled_data, row.names = 1, data.type = "integer", info.cols = c(2:5))

# Calculate WIC/TNW
WT_geno_subsamples = WTdMC(rinsp_geno_subsamples, replicates = 999, pop.diet = "average")
WT_geno_subsamples$WonT 
WT_geno_subsamples$p.value

# Calculate the proportional similarity index (PSi)
PSi_geno_subsamples = PSicalc(rinsp_geno_subsamples, pop.diet = "average", exclude = FALSE, replicates=999, precision = 1e-9)
PSi_geno_subsamples$IS
PSi_geno_subsamples$IS.pvalue
```


Conduct 1,000 iterations. This process takes a long time to run, so results are saved and stored under the file name: ______.

```{r}
# Set up vectors to store results
wonT_values <- numeric(100)
wonT_pvalues <- numeric(100)
IS_values <- numeric(100)
IS_pvalues <- numeric(100)

# Run the sampling and calculation 100 times
for (i in 1:100) {
  # Sample one scat per individual
  sampled_data <- geno_samples %>% 
    group_by(Individual) %>%
    sample_n(size = 1, replace = TRUE) %>%
    ungroup() %>% 
    mutate(ID = seq(1, nrow(.))) %>% 
    relocate(ID, .before = SampleID) %>% 
    as.data.frame()
  
  # Create RInSp object
  rinsp_geno_subsamples <- import.RInSp(sampled_data, row.names = 1, data.type = "integer",
                                        info.cols = c(2:5))
  
  # Calculate WIC/TNW
  WT_geno_subsamples <- WTdMC(rinsp_geno_subsamples, replicates = 999, pop.diet = "average",
                              print.ris = FALSE)
  wonT_values[i] <- WT_geno_subsamples$WonT
  wonT_pvalues[i] <- WT_geno_subsamples$p.value
  
  # Calculate the Proportional Similarity index (PSi)
  PSi_geno_subsamples <- PSicalc(rinsp_geno_subsamples, pop.diet = "average", 
                                 exclude = TRUE, replicates = 999, precision = 1e-9)
  IS_values[i] <- PSi_geno_subsamples$IS
  IS_pvalues[i] <- PSi_geno_subsamples$IS.pvalue
}

# Save model output
# saveRDS()

# Calculate mean and 95% credible intervals for WonT and IS
mean_wonT <- mean(wonT_values)
mean_IS <- mean(IS_values)
ci_wonT <- ci(wonT_values, ci = 0.95)
ci_IS <- ci(IS_values, ci = 0.95)

# Calculate the proportion of p-values < 0.05 for WonT and IS
prop_wonT_pval <- mean(wonT_pvalues < 0.05) ## IS THIS CORRECT?
prop_IS_pval <- mean(IS_pvalues < 0.05)

# Summary of results
summary_results <- data.frame(
  Metric = c("WonT", "IS"),
  Mean = c(mean_wonT, mean_IS),
  CI_lower = c(ci_wonT$CI_low, ci_IS$CI_low),
  CI_upper = c(ci_wonT$CI_high, ci_IS$CI_high),
  Pval_below_0.05 = c(prop_wonT_pval, prop_IS_pval))
```

## Integrative approach averaging across all samples per individual

To avoid throwing away data, an alternative approach involves utilizing all scats collected per individual. However, RInSp requires that each row represent an individual, not a sample. Accordingly, the code below calculates the average relative read abundance per species per individual. To transform these means into integers, the means are then multiplied by 1,000 and rounded to the nearest whole number. This ensures that integer data created reflect the RRA proportions ascribed to each individual. Once again, pop.diet = "average" is used to ensure that individual diets are first converted into proportions and only then are those proportions averaged for each resource to calculate the population-level diet.

```{R}
# Create data frame of averaged RRA values per individual and convert to integers
geno_RRA <- inner_join(metadata, sp.RRA, by = "SampleID") %>% 
  filter(Category == "SFCoy") %>% 
  group_by(Individual) %>% 
  summarize(across(where(is.numeric), ~ round(mean(.) * 1000)))  %>% # multiple by 1000 & round
  mutate(ID = seq(1, nrow(.))) %>% 
  relocate(ID, .before = Individual) %>% 
  as.data.frame()

# Create RInSp object
rinsp_geno_RRA = import.RInSp(geno_RRA, row.names = 1, data.type = "integer", info.cols = 2)

# WIC/TNW
WT_geno_RRA = WTdMC(rinsp_geno_RRA, replicates=999, print.ris=TRUE, pop.diet = "average")
WT_geno_RRA$WonT #0.475
WT_geno_RRA$p.value #0.001

# Proportional similarity index (PSi)
PSi_geno_RRA = PSicalc(rinsp_geno_RRA, pop.diet = "average", exclude = FALSE, replicates=999, precision = 1e-9)
PSi_geno_RRA$IS #0.49
PSi_geno_RRA$IS.pvalue #0.001
```

Estimates for WIC/TNW and PSi are very similar, at ~0.5. The problem with this approach is that many indidivduals only have a single sample, but other individuals have several samples. The more samples per individual, the more opportunities they have to get a more representative diet, making them look like generalists, so this approach is problematic. It is best to have equal(ish) representation among individuals, which in this case, is a single scat.

# Territory-level differences

Measure the extent of individual specialization within each territory.

```{r}
# Define sites to drop
sites.to.drop <- c("Stern Grove", "Sunset Res")

# Define a function to process each territory
process_territory <- function(territory_name) {
  # Filter, group, summarize, and format data
  geno_data <- inner_join(metadata, sp.RRA, by = "SampleID") %>%
    left_join(fams, by="Individual") %>% 
    filter(Category == "SFCoy", Family.Group == territory_name) %>%
    group_by(Individual) %>%
    summarize(across(where(is.numeric), ~ round(mean(.) * 1000))) %>%
    mutate(ID = seq(1, nrow(.))) %>%
    relocate(ID, .before = Individual) %>%
    as.data.frame()
  
  # Get count of number of individuals included per territory
  row_count <- nrow(geno_data)
  
  # Create RInSp object
  rinsp_obj <- import.RInSp(geno_data, row.names = 1, data.type = "integer", info.cols = 2)
  
  # Calculate WIC/TNW index
  WT_result <- WTdMC(rinsp_obj, replicates = 999, print.ris = TRUE, pop.diet = "average")
  
  # Return results for WonT and p-value
  list(Family.Group = territory_name, WonT = WT_result$WonT, p_value = WT_result$p.value,
       Num.Ind = row_count)
}

# Get unique territory names
unique_fams <- unique(fams$Family.Group)

# Apply the function to each territory and store results
results <- map(unique_fams, process_territory)

# Create data frame of averages for diet functional groups for each territory
diet_terr <- left_join(func.RRA, metadata, by="SampleID") %>% 
  group_by(Territory) %>% 
  summarize(Anthropogenic=mean(Anthropogenic),
            Small.Mammal=mean(Small.Mammal),
            Medium.Mammal=mean(Medium.Mammal),
            Birds = mean(Bird)) %>% 
  filter(!Territory %in% sites.to.drop)

# Convert list of results to a data frame and add covariate data
results_df <- bind_rows(results) %>% 
  mutate(BIC = 1-WonT, .after=WonT) %>% 
  filter(Family.Group != "Transient") %>% # drop transient individual
  left_join(covs, by = c("Family.Group" = "Territory")) %>%  # join environmental covariates
  left_join(diet_terr, by = c("Family.Group" = "Territory")) # join territory-level functional group diet data

# View the results
arrange(results_df, WonT)

# Make plots
results_df %>% 
  ggplot(aes(x=Small.Mammal*100, y=WonT))+
  geom_point()+
  geom_smooth(method="lm")+
  geom_hline(yintercept=0.475, linetype = "dashed")+
  geom_text(label=results_df$Family.Group)+
  theme_custom()+
  ylab("WIC/TNW")+xlab("Average RRA Small Mammals")

results_df %>% 
  ggplot(aes(x=Anthropogenic*100, y=WonT))+
  geom_point()+
  geom_smooth(method="lm")+
  geom_hline(yintercept=0.475, linetype = "dashed")+
  geom_text(label=results_df$Family.Group)+
  theme_custom()+
  ylab("WIC/TNW")+xlab("Average RRA Human Food")

results_df %>% 
  ggplot(aes(x=ISA, y=WonT))+
  geom_point()+
  geom_smooth(method="lm")+
  geom_hline(yintercept=0.475, linetype = "dashed")+
  geom_text(label=results_df$Family.Group)+
  theme_custom()+
  ylab("WIC/TNW")+xlab("Impervious Surface Area")

# Model
mod.WonT.ISA <- lm(WonT~ISA, data=results_df)
summary(mod.WonT.ISA)

mod.WonT.Anthro <- lm(WonT~Anthropogenic, data=results_df)
summary(mod.WonT.Anthro)

mod.WonT.SmallMam <- lm(WonT~Small.Mammal, data=results_df)
summary(mod.WonT.SmallMam)
```

# Individual data

Subsample 8 scats per individual. Do this 1,000 times and calculate an average specialization index with confidence interval

```{r}
# Create subsampled dataset with 8 scats per individual and raw sequence reads
ind_samples <- inner_join(metadata, sp.RRA, by = "SampleID") %>% 
  filter(Category == "SFCoy") %>% 
  group_by(Individual) %>%  
  mutate(count = n(), .after="Individual") %>% 
  filter(count >=8) %>%
  select(-c(count)) %>% 
  group_by(Individual) %>% 
  summarize(across(where(is.numeric) & !any_of("ID"), ~ round(mean(.) * 100000)))  %>% 
  mutate(ID = seq(1,nrow(.))) %>% 
  relocate(ID, .before=Individual) %>% 
  left_join(fams, by="Individual") %>% 
  relocate("Sex", .after="Individual") %>% 
  relocate("Family.Group", .after="Sex")

# Create RInSp object
rinsp_ind_RRA = import.RInSp(ind_samples, row.names = 1, data.type = "integer", info.cols = c(2:4))

# WIC/TNW relative to the whole population diet
WT_ind_RRA = WTdMC(rinsp_ind_RRA, replicates=999, print.ris=TRUE, pop.diet = "average")
WT_ind_RRA$WonT #0.727
WT_ind_RRA$p.value #0.001

# Proportional similarity index (PSi) relative to the whole population diet
PSi_ind_RRA = PSicalc(rinsp_ind_RRA, pop.diet = "average", exclude = FALSE, replicates=999, precision = 1e-9)
PSi_ind_RRA$IS #0.653
PSi_ind_RRA$IS.pvalue #0.001

PSi_inds_df <- data.frame(PSi = PSi_ind_RRA$PSi) %>% 
  mutate(ID = seq(1,nrow(.))) %>% 
  left_join(ind_samples %>% select(ID, Individual), by = "ID") %>% 
  left_join(fams, by="Individual") %>% 
  left_join(covs, by = c("Family.Group" = "Territory"))

# Calculating IS within each family group using the high-quality data
table(ind_samples$Family.Group)

# Subset dfs for each family group
inds_Bernal <- ind_samples %>% filter(Family.Group=="Bernal")
inds_Coit <- ind_samples %>% filter(Family.Group=="Coit")
inds_Corona <- ind_samples %>% filter(Family.Group=="Corona")
inds_Glen <- ind_samples %>% filter(Family.Group=="Glen")
inds_Merced <- ind_samples %>% filter(Family.Group=="Merced")
inds_Presidio <- ind_samples %>% filter(Family.Group=="Presidio")

# Create RInSp objects
rinsp_inds_Bernal = import.RInSp(inds_Bernal, row.names = 1, data.type = "integer", info.cols = c(2:4))
rinsp_inds_Coit = import.RInSp(inds_Coit, row.names = 1, data.type = "integer", info.cols = c(2:4))
rinsp_inds_Corona = import.RInSp(inds_Corona, row.names = 1, data.type = "integer", info.cols = c(2:4))
rinsp_inds_Glen = import.RInSp(inds_Glen, row.names = 1, data.type = "integer", info.cols = c(2:4))
rinsp_inds_Merced = import.RInSp(inds_Merced, row.names = 1, data.type = "integer", info.cols = c(2:4))
rinsp_inds_Presidio = import.RInSp(inds_Presidio, row.names = 1, data.type = "integer", info.cols = c(2:4))

# Calculate WIC/TNW index
WT_inds_Bernal <- WTdMC(rinsp_inds_Bernal, replicates = 999, print.ris = TRUE, pop.diet = "average")
WT_inds_Coit <- WTdMC(rinsp_inds_Coit, replicates = 999, print.ris = TRUE, pop.diet = "average")
WT_inds_Corona <- WTdMC(rinsp_inds_Corona, replicates = 999, print.ris = TRUE, pop.diet = "average")
WT_inds_Glen <- WTdMC(rinsp_inds_Glen, replicates = 999, print.ris = TRUE, pop.diet = "average")
WT_inds_Merced <- WTdMC(rinsp_inds_Merced, replicates = 999, print.ris = TRUE, pop.diet = "average")
WT_inds_Presidio <- WTdMC(rinsp_inds_Presidio, replicates = 999, print.ris = TRUE, pop.diet = "average")

# Combine the WonT values into a data frame
results_inds_df <- tibble(
  Territory = c("Bernal", "Coit", "Corona", "Glen", "Merced", "Presidio"),
  WonT = c(
    WT_inds_Bernal$WonT,
    WT_inds_Coit$WonT,
    WT_inds_Corona$WonT,
    WT_inds_Glen$WonT,
    WT_inds_Merced$WonT,
    WT_inds_Presidio$WonT)) %>% 
  left_join(covs, by="Territory")

ggplot(results_inds_df, aes(x=ISA, y=WonT))+
  geom_point()+
  geom_smooth(method="lm")+
  geom_text(label=results_inds_df$Territory)+
  geom_hline(yintercept=0.727, linetype="dashed")+
  ylim(.4,1.)+
  theme_custom()

# Calculate PSi's within each family group for each individual
PSi_inds_Bernal = PSicalc(rinsp_inds_Bernal, pop.diet = "average", exclude = FALSE, replicates=999, precision = 1e-9)
PSi_inds_Coit = PSicalc(rinsp_inds_Coit, pop.diet = "average", exclude = FALSE, replicates=999, precision = 1e-9)
PSi_inds_Corona = PSicalc(rinsp_inds_Corona, pop.diet = "average", exclude = FALSE, replicates=999, precision = 1e-9)
PSi_inds_Glen = PSicalc(rinsp_inds_Glen, pop.diet = "average", exclude = FALSE, replicates=999, precision = 1e-9)
PSi_inds_Merced = PSicalc(rinsp_inds_Merced, pop.diet = "average", exclude = FALSE, replicates=999, precision = 1e-9)
PSi_inds_Presidio = PSicalc(rinsp_inds_Presidio, pop.diet = "average", exclude = FALSE, replicates=999, precision = 1e-9)
```




1) Calculate specialization for the whole population - across the SF coyote population how much diet variation is attributed to among vs. within individual differences? (subsampling and integrative approaches

    * subsampling approach TBD but probably LOW
    * integrative approach = 0.475 (47.5% of variation attribute to WITHIN INDIVIDUAL VAR) ->         meaning that MORE of the variation is attributed to BETWEEN INDIVIDUAL VAR

2) Now we can zoom into our family groups: do some have more vs. less IS? YES - seems like anthro-eating groups have less evidence of IS then natural-eating groups (integrative approach; fine bc all groups biased similarly). Can I use these territory-level WIC/TNW values to show the IS is lower (more var explained by within-ind diffs) within family groups than across the pop?
  
    * Yes so most of the family groups have WIC/TNW values that are GREATER than the population     estimate of IS. This indicates to us that WITHIN FAMILIES there is not strong evidence of IS     but  

Now the question is whether or not these results are repeatable when we use the individual for which we have strong longitudinal sampling (multiple scats, 8 or more). When using this population of individuals, we find CONSIDERABLY less evidence for IS in this subset of individuals (0.73) BUT when we look at the family groups we have (which is only 6 with more than one individual) we see a similar pattern where the estimated amount of within-group IS is even less (WIC/TNW of 0.77-0.94).

What all if this points to is pretty moderate IS (esp. for a species like a coyote) across the population, but considerably less strong IS within family groups, with the least strong IS in the groups that have the least access to natural prey.

# Taking random subsamples and seeing what the WIC/TNW values are to see if my estimates are strongly different from random

```{r}
# let's take a totally random subsample of 100 scats
all_data <- inner_join(metadata, sp.RRA, by = "SampleID") %>% 
  filter(Category == "SFCoy") %>% 
  mutate(rand.ID = sample(1:111, size = n(), replace = TRUE), .after=Individual) %>% 
  group_by(rand.ID) %>%
  summarize(across(where(is.numeric), ~ round(mean(.) * 10000))) %>%
  mutate(ID = seq(1, nrow(.))) %>%
  relocate(ID, .before = rand.ID) %>%
  as.data.frame()


genotype_data <- inner_join(metadata, sp.RRA, by = "SampleID") %>% 
  filter(Category == "SFCoy") %>% 
  sample_n(size = 100, replace = FALSE) %>% 
  mutate(rand.ID = sample(1:15, size = n(), replace = TRUE), .after=Individual) %>% 
  group_by(rand.ID) %>%
  summarize(across(where(is.numeric), ~ round(mean(.) * 10000))) %>%
  mutate(ID = seq(1, nrow(.))) %>%
  relocate(ID, .before = rand.ID) %>%
  as.data.frame()
  
  
  # Create RInSp object
all_rand <- import.RInSp(all_data, row.names = 1, data.type = "integer", info.cols = 2)
rinsp_obj_temp <- import.RInSp(genotype_data, row.names = 1, data.type = "integer", info.cols = 2)
  
  # Calculate WIC/TNW index
WT_result_rand <- WTdMC(all_rand, replicates = 999, print.ris = TRUE, pop.diet = "average") # here all the IDs are totally scrambled - let's see if there still evidence of specialization or not; if not, i'll feel good that my estimate of ~0.48 is legit # RESULT: 0.6485188
WT_result_temp <- WTdMC(rinsp_obj_temp, replicates = 999, print.ris = TRUE, pop.diet = "average") # here all the ID's for a subset of 100 samples (a random Territory) are all scrambled. i'm doing this to see if i still get considerably less variation explained by WIC compared to the scrambled all data; if it's similar to the estimate above then I'll know that the reduced sample size is not what is causing a reduction in variance explained by BIC. RESULT: 0.71; so the diff between my pop random IS and my territory random IS is negligible. But in my actual data i see strong diffs between Bernal (>0.653) Coit (0.747) and the population IS which is 0.48

```



OKAY SO TALKED WITH BEN:
-only the subsampling approach really makes sense because otherwise im inflating the diversity for the freq sampled coyotes and that messes with IS estimates
-so sticking with the subsampling approach i will iterate 1000 times to estimate a pop mean and CI
-then i will do the subsampling approach at the TERRITORY level and calculate a mean WoT value for each territory (and CI??)
-then i will plot the MEANS for each territory as a function of ISA to hopefully show that: where coys have less access to natural prey, they will all use human food sources, which decreases specialization; but where they have access to natural prey, opportunities allow indiviudals to specialize and diversity their diets relative to one another; therefore, in a city which has interspersed more natural and less natural areas, we see a lot of indiviudal level variation and specialization as a function of the landscape.

-will need to figure out how to run the code for the territory level situations.











